---
title: " 02_bibliometrics_data_subsetting_clean"
author: "Mackenzie Johnson"
date: "20260401"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
library(tidyverse)
library(stringr)
```


```{r code for writing csvs}

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```



```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```


[UPDATE:  4/24/2026]
Changed tactic - realized that finding the "overlapping" references probably misses references to older works, so generated the list of UNIQUE references from both columns of "refs" (in Step 6) and then further filtered the "works" to just those unique refs (in Step 7).  

Step 1)  filter the "works" rds by:
  - language (keep only those works in English)
  - type (keep only articles and reviews)
  - is.retracted (keep only those that have NOT been retracted)
 
 Step 2) verify those worked by looking at the number of distinct values
 
 Step 3) filter out additional observations by:
  - publication year (keep only pre-2026)
  - title (remove those with blank titles)
  - primary topic (remove NA) [see code in "02b" for exploration of this variable]
  - fwci (remove NA) [see code in "02b" for exploration of this variable]
  - references_counted (remove any 0 values) [see code in "02b" for exploration of this variable]

Step 4) select out several variables
  - (In Liza's code, she only used work_id, publication year, language, type, and referenced_works_count, but I will keep other variables.)


```{r subsetting the "works" rds}

# originally had this filtered to just "article" but based on Wood et al. (2025), decided to also keep "review", "book" and "book-chapter" 

filt_works <- works %>% 
  filter(language == "en") %>% 
  filter(type == "article" | type == "review") %>%  # | type == "book" | type == "book-chapter") 
  filter(is_retracted == "FALSE") %>% 
  filter(publication_year != 2026) %>% 
  filter(title != "") %>% 
  filter(!is.na(primary_topic)) %>% 
  filter(!is.na(fwci)) %>% 
  filter(referenced_works_count != 0) %>% 
  select(-pmid, -mag, -publication_date, -language, -is_retracted, -is_paratext, -biblio_volume, -biblio_issue, -biblio_first_page, -biblio_last_page, -primary_source_type, -primary_source_issn_l, -created_date, -updated_date)


#  want to test to make sure the filtering worked

n_distinct(filt_works$language)  # [1] 1  IT WORKED
n_distinct(filt_works$type)  # [1] 2  IT WORKED
n_distinct(filt_works$is_retracted)  # [1] 1  IT WORKED
n_distinct(filt_works$is_paratext)  # [1] 1  Didn't filter by this variable, but expected this result

# now I can go back and select out more columns
# end result for now is 21 variables
    # could probably select out more, but want to look at more bibliometric analyses first

```


Moving on to the "refs" rds

Step 5) keep only those articles that were cited at least 5 times (per Wood et al. 2025)  

Step 6) Once I have filtered it down to only those references with 5 or more citations, generate a list of just the unique values from both the "citing" and "cited" columns (as in, it's unique to BOTH columns). This step is NEW - before, I was limiting the refs to just those IDs that appeared as both citing and cited, but that would selectively miss older works that are part of the conversation (older articles CAN'T cite newer ones).


```{r subsetting the "refs" rds - unique}

refs_citing_only_v2 <- refs %>% 
  count(citing_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, citation = citing_id)

n_distinct(refs_citing_only_v2)  # [1] 43482
  
refs_cited_only_v2 <- refs %>% 
  count(cited_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, citation = cited_id)

n_distinct(refs_cited_only_v2)  # [1] 33024

nrow(refs_cited_only_v2) + nrow(refs_citing_only_v2)  # [1] 76506

refs_filt_unique <- bind_rows(refs_citing_only_v2, refs_cited_only_v2) %>% 
  unique()

```


Step 7) Limit the "filt_works" to just those unique references in "refs"


```{r subsetting the filtered works to match the unique refs}

filt_unique_works <- filt_works %>% 
  filter(work_id %in% refs_filt_unique$citation)

nrow(filt_unique_works) # [1] 54917
nrow(refs_filt_unique) - nrow(filt_unique_works)  # [1] 1634

print(count(filt_unique_works, is.na(work_id)))  # 1:          FALSE 54917, no so "NA" values

# so, there are 1634 references in the overlapping "refs" that 
#   do not show up when I intersect the filtered works with these overlapping references

# see "02b" for the process to identify the 1634
# all have been removed during the filtering process
# therefore, can use the object "filt_unique_works" for subsequent analysis

# will save this as the nodelist for network analysis

saveRDS(filt_unique_works, "network_nodelist")

```


Step 8) Considered filtering based on the different "primary_topic*" variables.  Did so both before and after I used %in% to limit the works to just those appearing in the overlapping reference list.  [see code in "02b" for exploration of these variables].  Ultimately decided NOT to filter on this topic for overall analysis, but could see diving deeper into some of the primary topic fields. 

[UPDATE]  

Step 9) As part of considering whether or not to filter based on the different "primary_topic*" variables, I also wanted to assess the distribution of "Indigenous" specific language in both the title and primary topic.  Takeaway:  In the filtered works that appear in the overlapping list of references, there are 56 observations that have "Indigenous|indigenous|tribal|tribe|Tribe|Native") in either the title or primary topic field - which equates to 0.28% of the works.  (In the full list of filtered references, including those not found in the overlapping references, there are 2860 observations - which equates to 2.8%.)  [see code in "02b" for exploration of this occurrence] 

Step 10) Having created the node list for the network analysis, need to also create the edge list.

```{r Creating the edgelist}

network_edgelist_full <- refs %>% 
  filter(citing_id %in% refs_citing_only$from | 
           cited_id %in% refs_cited_only$to)

network_edgelist_overlap1 <- refs %>% 
  filter(citing_id %in% filt_overlap_works$work_id | 
                       cited_id %in% filt_overlap_works$work_id)

# n = 460956

# remove self references

network_edgelist_overlap2 <-network_edgelist_overlap1[citing_id != cited_id]

saveRDS(network_edgelist_overlap2, "network_edgelist_reduced")

```
