---
title: " 02_bibliometrics_data_subsetting"
author: "Mackenzie Johnson"
date: "2026-03-07"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
library(tidyverse)
```

```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```

Looking over Wood et al. (2025) (Liza, Francesca, Tyler - bibliometric analysis of "environmental governance processes"), it's obviously acceptable for me to limit the analysis to just English-language works.  But even before I get to that, I want to subset out the data by type.  Recall, there are 12 types of works included.  When I first looked at the ones marked "books" for [type], I noticed that all the ones from American Indian Quarterly are book * reviews * , not books (!) - and the first non-AIQ entry * is* a book, but it's assigned a DOI that actually matches another article in the list of works.  

Maybe I should be using the [primary_source_type] instead - NOT!  All 146k+ are listed as "journal" for that variable.



```{r subset based on "type" of work}

works_books <- works %>% 
  filter(type == "book") # %>%   select(work_id, type, title, doi)

works_datasets <- works %>% 
  filter(type == "dataset") # %>%   select(work_id, type, title, doi)

n_distinct(works$primary_source_type)  
# [1] 1 - which means they're all "journal" ... 
# so I can easily SELECT out this variable ...
# and it also means if I'm going to subset based on type of work, it has to be done using the [type] variable

works_errata <- works %>% 
  filter(type == "erratum") # %>%   select(work_id, type, title, doi)



```





```{r limit to English-language}



```





```{r Data Exploration - "works" - publication dates }


```



 

```{r Data Exploration - "works" - filter potential 1 }


```




```{r Data Exploration - "works" - filter potential 2 }


```

