---
title: " 02_bibliometrics_data_subsetting"
author: "Mackenzie Johnson"
date: "20260307_11"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
library(tidyverse)
```

```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```

Looking over Wood et al. (2025) (Liza, Francesca, Tyler - bibliometric analysis of "environmental governance processes"), it's obviously acceptable for me to limit the analysis to just English-language works.  But even before I get to that, I want to subset out the data by type.  Recall, there are 12 types of works included.  When I first looked at the ones marked "books" for [type], I noticed that all the ones from American Indian Quarterly are book * reviews * , not books (!) - and the first non-AIQ entry * is* a book, but it's assigned a DOI that actually matches another article in the list of works.  

Maybe I should be using the [primary_source_type] instead - NOT!  All 146k+ are listed as "journal" for that variable.



```{r subsetting the "works" rds}

works_books <- works %>% 
  filter(type == "book") # %>%   select(work_id, type, title, doi)

works_datasets <- works %>% 
  filter(type == "dataset") # %>%   select(work_id, type, title, doi)

# n_distinct(works$primary_source_type)  
# [1] 1 - which means they're all "journal" ... 
# so I can easily SELECT out this variable ...
# and it also means if I'm going to subset based on type of work, it has to be done using the [type] variable

works_errata <- works %>% 
  filter(type == "erratum") # %>%   select(work_id, type, title, doi)

filtered_works <- works %>% 
  filter(language == "en") %>% 
  filter(type == "article") %>% 
  filter(is_retracted == "FALSE") %>% 
  select(-pmid, -mag, -primary_source_issn_l, -language, -type, -is_retracted, -is_paratext, -primary_source_type, -created_date, -updated_date)

#  want to test to make sure the filtering worked
n_distinct(filtered_works$language)  # [1] 1  IT WORKED
n_distinct(filtered_works$type)  # [1] 1  IT WORKED
n_distinct(filtered_works$is_retracted)  # [1] 1  IT WORKED
n_distinct(filtered_works$is_paratext)  # [1] 1  Didn't filter by this, but this is the expected result

# now I can go back and select out more columns
# end result for now is 25 variables
    # could probably select out more, but want to look at more bibliometric analyses first

```


Now want to try working with the "refs.rds" file.  1) I want to figure out how to get just the articles that appear in both columns.  Looked online: will try INTERSECT function first.  2) I want to somehow filter it by just those articles that were cited at least 5 times (per Wood et al. 2025).  


```{r subsetting the "works" rds}

# First, try getting a dataframe of just those articles that appear in both columns (citing and cited)

refs_overlap <- refs %>% 
  intersect(citing_id, cited_id)

  # Error in `intersect()`:
  # ! `...` must be empty.
  # ✖ Problematic argument:
  # • ..1 = cited_id
  # ℹ Did you forget to name an argument?
  # Run `rlang::last_trace()` to see where the error occurred.

refs_citing <- refs %>% 
  select(-cited_id)

refs_cited <- refs %>% 
  select(-citing_id)

refs_overlap <- intersect(refs_citing, refs_cited)

  # Error in `intersect()`:
  # ! `x` and `y` are not compatible.
  # ✖ Cols in `y` but not `x`: `cited_id`.
  # ✖ Cols in `x` but not `y`: `citing_id`.
  # Run `rlang::last_trace()` to see where the error occurred.

# ?intersect
  # column names have to be the same!!

# refs_citing_renamed <- refs_citing %>% 
  # rename(citing_id = citation)

  # Error in `rename()`:
  # ! Can't rename columns that don't exist.
  # ✖ Column `citation` doesn't exist.
  # Run `rlang::last_trace()` to see where the error occurred.


refs_citing <- refs %>% 
  select(-cited_id, citation = citing_id)

refs_cited <- refs %>% 
  select(-citing_id, citation = cited_id)

refs_overlap <- intersect(refs_citing, refs_cited)
typeof(refs_overlap) # [1] "list"
class(refs_overlap)  # [1] "data.table" "data.frame"

n_distinct(refs_overlap$citation)  # [1] 60584

# um...that's surprising
# need to back up - it worked as expected, giving me a df with one row for each of the ids found in both the citing and cited

# tally_refs_overlap <- count(refs_overlap, citation, sort = TRUE, name = "tally")
# min(tally_refs_overlap$tally)
# max(tally_refs_overlap$tally)

refs_citing_grouped <- refs_citing %>% 
  group_by(citation)

length(refs_citing_grouped$citation)  # [1] 602219

# ok, so none of the citing_ids are repeated?
# that can't possibly be right:

n_distinct(refs_citing_grouped$citation)  # [1] 79643

refs_cited_grouped <- refs_cited %>% 
  group_by(citation)

length(refs_cited_grouped$citation)  # # [1] 602219


# and none of the cited IDs are repeated ??
# that can't possibly be right!  

n_distinct(refs_cited_grouped$citation)  # [1] 73405

```





```{r Data Exploration - "works" - publication dates }


```



 

```{r Data Exploration - "works" - filter potential 1 }


```




```{r Data Exploration - "works" - filter potential 2 }


```

