---
title: " 02_bibliometrics_data_subsetting_clean"
author: "Mackenzie Johnson"
date: "20260401"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
library(tidyverse)
library(stringr)
```


```{r code for writing csvs}

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```



```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```

Step 1)  filter the "works" rds by:
  - language (keep only those works in English)
  - type (keep only articles and reviews)
  - is.retracted (keep only those that have NOT been retracted)
 
 Step 2) verify those worked by looking at the number of distinct values
 
 Step 3) filter out additional observations by:
  - publication year (keep only pre-2026)
  - title (remove those with blank titles)
  - primary topic (remove NA)
  - fwci (remove NA)
  - references_counted (remove any 0 values) [see code chunk in "02b" for exploration of this occurrence]

Step 4) select out several variables
  - (In Liza's code, she only used work_id, publication year, language, type, and referenced_works_count, but I will keep other variables.)


```{r subsetting the "works" rds}

# originally had this filtered to just "article" but based on Wood et al. (2025), decided to also keep "review", "book" and "book-chapter" 

filt_works <- works %>% 
  filter(language == "en") %>% 
  filter(type == "article" | type == "review") %>%  # | type == "book" | type == "book-chapter") 
  filter(is_retracted == "FALSE") %>% 
  filter(publication_year != 2026) %>% 
  filter(title != "") %>% 
  filter(!is.na(primary_topic)) %>% 
  filter(!is.na(fwci)) %>% 
  filter(referenced_works_count != 0) %>% 
  select(-pmid, -mag, -publication_date, -language, -is_retracted, -is_paratext, -biblio_volume, -biblio_issue, -biblio_first_page, -biblio_last_page, -primary_source_type, -primary_source_issn_l, -created_date, -updated_date)


#  want to test to make sure the filtering worked

n_distinct(filt_works$language)  # [1] 1  IT WORKED
n_distinct(filt_works$type)  # [1] 2  IT WORKED
n_distinct(filt_works$is_retracted)  # [1] 1  IT WORKED
n_distinct(filt_works$is_paratext)  # [1] 1  Didn't filter by this variable, but expected this result

# now I can go back and select out more columns
# end result for now is 21 variables
    # could probably select out more, but want to look at more bibliometric analyses first


print(count(filt_works, is.na(title)))  # 1:        FALSE 142103 -- as in, all of them, so no NA values

print(count(filt_works, is.na(fwci)))  # 2:        TRUE  10694
print(count(filt_works, is.na(primary_topic)))  # 2: TRUE   10680  
print(count(filt_works, is.na(primary_topic_subfield)))  # 2: TRUE   10680
print(count(filt_works, is.na(primary_topic_field)))  # 2: TRUE   10680
print(count(filt_works, is.na(primary_topic_domain)))  # 2: TRUE   10680

#  after filtering out the NA values for "primary_topic" - all the "primary_topic_..." variables have no NA values
# and "fwci" only has 14 values, so I'll filter those out 
# can see from Global Environments it worked - now there are 131409 observations

```


Moving on to the "refs" rds

Step 5) keep only those articles that were cited at least 5 times (per Wood et al. 2025)  

Step 6) Once I have filtered it down to only those references with 5 or more citations, find just those articles that appear as both "citing" and "cited" - use %in%


```{r subsetting the "refs" rds}

refs_citing_only <- refs %>% 
  count(citing_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, from = citing_id)
  
refs_cited_only <- refs %>% 
  count(cited_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, to = cited_id)


refs_filt_overlap <- refs_citing_only %>% 
  filter(from %in% refs_cited_only$to) %>% 
  mutate(citation_id = from) %>% 
  select(-from)

# cited_in_citing <- refs_cited_only %>% 
 #  filter(to %in% refs_citing_only$from)  

# also 19955 observations


n_distinct(refs_filt_overlap$citation_id)  # [1] 19955 - this matches the length of the dataframe!

# Seems to have worked  
# given that the number of distinct values equals the number of observations

```

Step 7) Filter out the rows with no references

```{r further subsetting the "works"}

# see code chunk in "02b" for exploration of this occurrence


filt_works <- filt_works %>% 
  filter(referenced_works_count != 0)


```


Step 8) Filter out rows based on primary topic information



Step 9) Next, limit the "works" entries to just those found to overlap in the "refs"




```{r investigating the "primary topic" variables of the filtered works}

# tally_primary_topic_field_filtered <- count(filt_works, primary_topic_field, sort = TRUE, name = "tally")
# str(tally_primary_topic_field_filtered)  # Classes ‘data.table’ and 'data.frame':	26 obs. of  2 variables

n_distinct(filt_works$primary_topic)  # [1] 2075
n_distinct(filt_works$primary_topic_domain)  # [1] 4
n_distinct(filt_works$primary_topic_field)  # [1] 26
n_distinct(filt_works$primary_topic_subfield)  # [1] 199

n_distinct(filt_works$primary_topic_subfield[filt_works$primary_topic_field == "Social Sciences"])  # [1] 21 (Wait, really?!?)
n_distinct(filt_works$primary_topic[filt_works$primary_topic_field == "Social Sciences"])  # [1] 535

tally_by_primary_field <- filt_works %>% 
  group_by(primary_topic_field) %>%
  # mutate(n = summarise(n(primary_topic))) %>%  # This does NOT work
  mutate(n_subfields = n_distinct(primary_topic_subfield)) %>% 
  mutate(n_topics = n_distinct(primary_topic)) %>% 
  select(primary_topic_field, n_subfields, n_topics) %>% 
  distinct(primary_topic_field, .keep_all = TRUE) # %>% 
  # arrange(desc(n))
  
indig_prim_field <- filt_works %>% 
  # group_by(primary_topic_field) %>%
  filter(str_detect(primary_topic, "Indigenous|indigenous|tribal|tribe|Tribe|Native")) %>% 
  group_by(primary_topic) %>%  # group_by(primary_topic_field)
  summarise(n_Indig = n())

indig_title <- filt_works %>% 
  # group_by(primary_topic_field) %>%
  filter(str_detect(title, "Indigenous|indigenous|tribal|tribe|Tribe|Native")) %>% 
  group_by(primary_topic_field) %>%  # group_by(primary_topic_field)
  summarise(n_Indig = n())
  
  output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(tally_by_primary_field, file = full_output_path_fields)

prim_field_biochem <- filt_works %>% 
  filter(primary_topic_field == "Biochemistry, Genetics and Molecular Biology")

prim_field_socsci <- filt_works %>% 
  filter(primary_topic_field == "Social Sciences")

indig_socsci <- str_count(prim_field_socsci$primary_topic, "Indig") %>% 
  filter()

prim_field_envsci <- filt_works %>% 
  filter(primary_topic_field == "Environmental Science")

prim_field_econ <- filt_works %>% 
  filter(primary_topic_field == "Economics, Econometrics and Finance")

prim_field_agbio <- filt_works %>% 
  filter(primary_topic_field == "Agricultural and Biological Sciences")


tally_biochem <- prim_field_biochem %>% 
  count(source_group, sort = TRUE, name = "tally") %>% 
  mutate(pct = tally / nrow(prim_field_biochem))

biochem_topics <- prim_field_biochem %>% 
  group_by(primary_topic_subfield) %>% 
  count(primary_topic, sort = TRUE, name = "n_topics") %>% 
  mutate(n_subfields = count(primary_topic_subfield, ))



```




```{r Elise's help}

# Elise reminded me of a simpler way to get the counts - use "table"

  # but to write the table to csv, need to make it a dataframe
  # as.data.frame(table(prim_field_biochem$primary_topic_subfield))

# just need to write them to a csv

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name <- "AgBio_subfields.csv"
full_output_path_fields <- file.path(output_directory, file_name)

write_excel_csv(as.data.frame(table(prim_field_agbio$primary_topic_subfield)), file = full_output_path_fields)

table(filt_works$primary_topic)
table(prim_field_biochem$primary_topic)
table(prim_field_biochem$primary_topic_subfield)

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```

```{r Elise's help} 

table(filt_works$)

```



```{r subsetting the filtered works to match the overlapping refs}

# lots of options
# start with "%in%" and then test the different join functions

test <- filtered_works %>% 
  filter(work_id %in% filtered_refs_overlap$citation)

print(count(test, is.na(work_id)))  # 1:          FALSE 18924 -- so no NA values, and 53 fewer observations than the refs alone

f_refs_overlap <- filtered_refs_overlap %>% 
  select(work_id = citation)

test2 <- inner_join(filtered_works, f_refs_overlap, by = "work_id")
  # Error in 'inner_join()'
  # `by` must be a (named) character vector, list, `join_by()` result, or NULL, not a function.
      # added "" around [citation]
        # Error in `inner_join()`:
        # ! Join columns in `x` must be present in the data.


  # typeof(filtered_refs_overlap$citation)

test3 <- left_join(filtered_works, f_refs_overlap, by = "work_id")
print(count(test3, is.na(work_id)))  # 1:          FALSE 126737 -- should have returned (126737-18977 = 107760) NA values...?
  # WAIT!  Shouldn't be looking for NA in the work_id, should be looking in another column - maybe "source_id"

test4 <- right_join(filtered_works, f_refs_overlap, by = "work_id")
print(count(test4, is.na(work_id)))  # 1:          FALSE 18977 -- good, would have been off if there were any NA values

test5 <- full_join(filtered_works, f_refs_overlap, by = "work_id")
print(count(test5, is.na(work_id)))  # 1:          FALSE 126790  --  wait, that's more than in "filtered_works" - by 53

# OK, so there are 53 observations that I'm filtering out of the "works" rds that 
# are still captured by my "filtered_refs_overlap" dataframe

```

```{r need to find the 53 citations from the overlapping references that aren't in the filtered works}

test5_na <- test5 %>%
  filter(is.na(source_id))

# I want to explore these manually since there are only 53 of them
# need to write a csv and do this in Excel

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"

# great, but I want it in the "data_output" folder

# setwd("~/data_output")  # Error in setwd("~/data_output") : cannot change working directory

# Googled it 
# Specify directory and file-name components, use them to construct the full path

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name <- "Refs_not_in_Works.csv"
full_output_path <- file.path(output_directory, file_name)

# write_excel_csv(test5_na, "Refs_not_in_Works")  # , "/data_output")

# write_excel_csv(test5_na, file = full_output_path)


##  ACTUALLY -- there's a faster way to do this...

refs_not_in_works <- works %>%
  filter(work_id %in% test5_na$work_id)

write_excel_csv(refs_not_in_works, file = full_output_path)

# OK, found them!  
# in the spreadsheet, I highlighted why each would have been filtered out
# so, should definitely use either the %in% or one of the joins that does NOT keep the "refs" that aren't in the "works"

```


```{r subsetting the "refs" rds using %in% }

refs_citing_only_2 <- refs %>% 
  count(citing_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, from = citing_id)
  
refs_cited_only_2 <- refs %>% 
  count(cited_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, to = cited_id)


citing_in_cited <- refs_citing_only_2 %>% 
  filter(from %in% refs_cited_only_2$to)

cited_in_citing <- refs_cited_only_2 %>% 
  filter(to %in% refs_citing_only_2$from)


# now that I have isolated both the "citing_id" works and the "cited_id" works into their own dataframes, 
# filtered them to just those works that appear 5 or more times
# and have renamed them with the same column name
# I should be able to intersect them

# refs_filtered_overlap_2 <- intersect(refs_citing_only_2, refs_cited_only_2)  # doesn't work b/c columns don't have same name

typeof(filtered_refs_overlap) # [1] "list"
class(filtered_refs_overlap)  # [1] "data.table" "data.frame"

n_distinct(refs_filtered_overlap$citation)  # [1] 18977 - this matches the length of the dataframe!

# um - I think it worked!  
# given that the number of distinct values equals the number of observations
# and given the below
    # tally_filtered_refs_overlap <- count(filtered_refs_overlap, citation, sort = TRUE, name = "tally")
    # min(tally_filtered_refs_overlap$tally)  # [1] 1
    # max(tally_filtered_refs_overlap$tally)  # [1] 1

```



```{r Data Exploration - "filt_works" - works with no references }

print(min(filt_works$referenced_works_count))
print(count(filt_works, referenced_works_count == 0))

    # referenced_works_count == 0      n
                            # <lgcl>  <int>
    # 1:                       FALSE 103075
    # 2:                        TRUE  28334  

count_no_refs <- as.integer(count(filt_works[referenced_works_count == 0]))  

tally_no_refs_by_source_group <- filt_works %>% 
  filter(referenced_works_count == 0) %>% 
  group_by(source_group) %>%  # group_by(source_id) -- gets the same tally as "by_journal" but now with source_id, not journal name
  summarise(n = n()) %>% 
  mutate(pct = n/count_no_refs) %>% 
  print()


tally_no_refs_by_journal <- filt_works %>% 
  filter(referenced_works_count == 0) %>% 
  group_by(primary_source_display_name) %>% 
  summarise(n = n()) %>% 
  mutate(pct = n/count_no_refs) %>% 
  arrange(desc(n))
  

count(referenced_works_count, sort = FALSE, name = "tally")

length(filt_works$referenced_works_count[filt_works$referenced_works_count < 10])  # 43782
length(filt_works$referenced_works_count[filt_works$referenced_works_count >= 10])  # 87627

no_refs_in_works <- filt_works %>% 
  filter(referenced_works_count == 0)

str(no_refs_in_works)  # Classes ‘data.table’ and 'data.frame':	25529 obs. of  21 variables

no_refs_overlap <- no_refs_in_works %>% 
  filter(work_id %in% refs_filtered_overlap$citation)

str(no_refs_overlap)  # Classes ‘data.table’ and 'data.frame':	0 obs. of  21 variables

no_refs_in_works_open <- filt_works %>% 
  filter(referenced_works_count == 0) %>% 
  filter(oa_status != "closed")

str(no_refs_in_works_open)  # Classes ‘data.table’ and 'data.frame':	2422 obs. of  21 variables
length(no_refs_in_works_open$has_fulltext[no_refs_in_works_open$has_fulltext == "TRUE"])  # 720 - these have full text
length(no_refs_in_works_open$oa_is_oa[no_refs_in_works_open$oa_is_oa == "TRUE"]) # [1] 2422 - good, SHOULD match the length of the df
n_distinct(no_refs_in_works_open$oa_is_oa)  # [1] 1 - yep, apparently they are all OA - that 1 value is "TRUE")


# OK, so seems like none of the references with zero values for the "referenced_works_count" show up in the "refs" rds
# even if their OA status is NOT closed


one_refs_in_works <- filtered_works %>% 
  filter(referenced_works_count == 1)

str(one_refs_in_works)  # Classes ‘data.table’ and 'data.frame':	2358 obs. of  21 variables

one_refs_overlap <- one_refs_in_works %>% 
  filter(work_id %in% refs$cited_id)

str(one_refs_overlap)  # Classes ‘data.table’ and 'data.frame':	1005 obs. of  21 variables

# BUT, as soon as there is a single referenced work, then there is overlap

```