---
title: " 02_bibliometrics_data_subsetting"
author: "Mackenzie Johnson"
date: "20260307_11"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
library(tidyverse)
```

```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```

Looking over Wood et al. (2025) (Liza, Francesca, Tyler - bibliometric analysis of "environmental governance processes"), it's obviously acceptable for me to limit the analysis to just English-language works.  But even before I get to that, I want to subset out the data by type.  Recall, there are 12 types of works included.  When I first looked at the ones marked "books" for [type], I noticed that all the ones from American Indian Quarterly are book * reviews * , not books (!) - and the first non-AIQ entry * is* a book, but it's assigned a DOI that actually matches another article in the list of works.  

(Maybe I should be using the [primary_source_type] instead - NOT!  All 146k+ are listed as "journal" for that variable.)

I can also limit the data to those articles that were NOT retracted.  


```{r subsetting the "works" rds}

# originally had this filtered to just "articles" but based on Wood et al. (2025), decided to also keep "reviews", "book" and "book-chapter" 

filtered_works <- works %>% 
  filter(language == "en") %>% 
  filter(type == "article" | type == "review" | type == "book" | type == "book-chapter") %>% 
  filter(is_retracted == "FALSE") %>% 
  select(-pmid, -mag, -publication_date, -language, -is_retracted, -is_paratext, -biblio_volume, -biblio_issue, -biblio_first_page, -biblio_last_page, -primary_source_type, -primary_source_issn_l, -created_date, -updated_date)


#  want to test to make sure the filtering worked

n_distinct(filtered_works$language)  # [1] 1  IT WORKED
n_distinct(filtered_works$type)  # [1] 4  IT WORKED
n_distinct(filtered_works$is_retracted)  # [1] 1  IT WORKED
n_distinct(filtered_works$is_paratext)  # [1] 1  Didn't filter by this variable, but expected this result

# now I can go back and select out more columns
# end result for now is 25 variables
    # could probably select out more, but want to look at more bibliometric analyses first

# maybe I should filter out anything from 2026...?
# print(count(filtered_works, publication_year == 2026))
    # publication_year == 2026      n
    # 1                    FALSE 137488
    # 2                     TRUE   1120  
    
# so if I filter out those, the "filtered_works" should have 137,488 observations

filtered_works <- filtered_works %>% 
  filter(publication_year != 2026) # 

# can see from the data in the Global Environment that it worked

# also want to filter out anything without a title
print(count(filtered_works, title == ""))
    # title == ""      n
    # 1       FALSE 137275
    # 2        TRUE    213


filtered_works <- filtered_works %>% 
  filter(title != "") 

# can see from the data in the Global Environment that it worked


print(count(filtered_works, is.na(title)))  # 1:        FALSE 137275 -- as in, all of them, so no NA values

print(count(filtered_works, is.na(fwci)))  # 2:        TRUE  10534
print(count(filtered_works, is.na(primary_topic)))  # 2: TRUE   10529  
print(count(filtered_works, is.na(primary_topic_subfield)))  # 2: TRUE   10529
print(count(filtered_works, is.na(primary_topic_field)))  # 2: TRUE   10529
print(count(filtered_works, is.na(primary_topic_domain)))  # 2: TRUE   10529

#  after filtering out the NA values for "primary_topic" - all the "primary_topic_..." variables have no NA values
# and "fwci" only has 5 values, so I'll filter those out below


filtered_works <- filtered_works %>% 
  filter(!is.na(primary_topic)) %>% 
  filter(!is.na(fwci))

# can see from the data in the Global Environment that it worked



```


Now want to try working with the "refs.rds" file.  1) I want to somehow filter it to just those articles that were cited at least 5 times (per Wood et al. 2025)  

2) Once I have filtered it down to only those references with 5 or more citations, I want to figure out how to get just the articles that appear in both columns.  Looked online: will try INTERSECT function first.  For this, I already made some progress in the "practice" file - see comments in the code chunk below. 


```{r subsetting the "refs" rds}

refs_citing_only <- refs %>% 
  count(citing_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, citation = citing_id)
  
refs_cited_only <- refs %>% 
  count(cited_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, citation = cited_id)


# now that I have isolated both the "citing_id" works and the "cited_id" works into their own dataframes, 
# filtered them to just those works that appear 5 or more times
# and have renamed them with the same column name
# I should be able to intersect them

filtered_refs_overlap <- intersect(refs_citing_only, refs_cited_only)

typeof(filtered_refs_overlap) # [1] "list"
class(filtered_refs_overlap)  # [1] "data.table" "data.frame"

n_distinct(filtered_refs_overlap$citation)  # [1] 18977 - this matches the length of the dataframe!

# um - I think it worked!  
# given that the number of distinct values equals the number of observations
# and given the below
    # tally_filtered_refs_overlap <- count(filtered_refs_overlap, citation, sort = TRUE, name = "tally")
    # min(tally_filtered_refs_overlap$tally)  # [1] 1
    # max(tally_filtered_refs_overlap$tally)  # [1] 1

```





 

```{r Data Exploration - "filtered_works" - works with no references }

print(min(filtered_works$referenced_works_count))
print(count(filtered_works, referenced_works_count == 0))

    # referenced_works_count == 0      n
                            # <lgcl>  <int>
    # 1:                       FALSE 101209
    # 2:                        TRUE  25529  # this is without the "Boraspati" works included

tally_no_refs_by_source_group <- filtered_works %>% 
  filter(referenced_works_count == 0) %>% 
  group_by(source_group) %>%  # group_by(source_id) -- gets the same tally as "by_journal" but now with source group, not journal name
  count(source_group, sort = TRUE, name = "tally") %>% 
  mutate(percent = tally/count_no_refs) %>% 
  print()

count_no_refs <- 25529

tally_no_refs_by_journal <- filtered_works %>% 
  filter(referenced_works_count == 0) %>% 
  group_by(source_id) %>%  # group_by(source_id) 
  count(primary_source_display_name, sort = TRUE, name = "tally") %>% 
  print()

# Wait, still have the "Boraspati" references

filtered_works <- filtered_works %>% 
  filter(primary_source_display_name != "Boraspati Journal Journal of Bilingualism Organization Research Articles Studies in Pedagogy Anthropology Theory and Indigenous Cultures")

test <- filtered_works %>% 
  filter(work_id %in% filtered_refs_overlap$citation)

print(count(test, is.na(work_id)))  # 1:          FALSE 18924 -- so no NA values, and 53 fewer observations than the refs alone

f_refs_overlap <- filtered_refs_overlap %>% 
  select(work_id = citation)

test2 <- inner_join(filtered_works, f_refs_overlap, by = "work_id")
  # Error in 'inner_join()'
  # `by` must be a (named) character vector, list, `join_by()` result, or NULL, not a function.
      # added "" around [citation]
        # Error in `inner_join()`:
        # ! Join columns in `x` must be present in the data.


  # typeof(filtered_refs_overlap$citation)

test3 <- left_join(filtered_works, f_refs_overlap, by = "work_id")
print(count(test3, is.na(work_id)))  # 1:          FALSE 126737 -- should have returned (126737-18977 = 107760) NA values...?

test4 <- right_join(filtered_works, f_refs_overlap, by = "work_id")
print(count(test4, is.na(work_id)))  # 1:          FALSE 18977 -- good, would have been off if there were any NA values

test5 <- full_join(filtered_works, f_refs_overlap, by = "work_id")
print(count(test5, is.na(work_id)))  # 1:          FALSE 126790  --  wait, that's more than in "filtered_works" - by 53

# OK, so there are 53 observations that I'm filtering out of the "works" rds that 
# are still captured by my "filtered_refs_overlap" dataframe

```




```{r Data Exploration - "works" - filter potential 2 }


```

