---
title: " 02_bibliometrics_data_subsetting_soc_env_sci"
author: "Mackenzie Johnson"
date: "20260430"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}

# library(ggplot2)

library(knitr)
library(tidyverse)
library(stringr)
```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```

[UPDATE:  5/12/2026

Because I am interested in seeing who all is citing NAIS-themed works and who NAIS-themed works are citing, I found the edgelist by filtering "refs" down to any rows that had one of the NAIS-themed works (based on title or primary_topic) in EITHER the citing or cited columns.  

BUT, that means my nodelist needs to include the works that are NOT in my NAIS-themed subset because - basically need my nodelist to match the edgelist.  So added "6)" to the workflow below.]

[UPDATE:  5/05/2026

Still ran into problems using the new edgelist in the network-generation stage.  Came back to this stage to generate the relevant "works" and "refs" dataframes that are limited to just those works with an NAIS-relevant keyword in either the title field or the primary_topic field.  

Also realized I need to switch the order in which I filter subset the refs.  Before I was removing the self-references and loops AFTER I limited the refs to just those IDs that appear 5 or more times, but actually I need to switch the order.  

New workflow for this stage:
1) filter the works
2) clean the refs (remove self-refs and loops)
3) generate a unique list of refs that have at least 5 occurrences
4) subset the works to just those unique refs (%in%)
5) subset the refs to match those works (%in%)]
6) make the nodelist match the edgelist

[UPDATE:  4/30/2026

After switching the approach to limit the 'works' and 'refs' to those UNIQUE refs, I tried using the resultant edgelist for the Workflow Stage 3 (networks) - but it didn't go well.  There were too many clusters, the main cluster still had 90k+ components, and the community detection algorithm did not produce usable results.  Thus, I want to filter both (the 'works' and the 'refs') to just examine the works that have a "primary_topic_field" value of "Social Sciences" and/or "Environmental Science" in order to derive a network/clusters that I can actually analyze.

Did NOT venture into step 10 (focus on NAIS-relevant topics/titles) today - want to test out the new edgelist in the next Workflow stage.]


[UPDATE:  4/24-27/2026]
Changed tactic - realized that finding the "overlapping" references probably misses references to older works, so generated the list of UNIQUE references (from both columns of "refs") (in Step 6) and then further filtered my "works_filt" to just those unique refs (in Step 7).  Created a new nodelist from this filtered works list of just those unique "refs".  Added an update to Step 8, which I might revisit in the future.  Edited the commentary for Step 9 because I added terms to the str_detect.  Then created a new nodelist in Step 10.  

----------------------

Step 1)  filter the "works" rds by:
  - language (keep only those works in English)
  - type (keep only articles and reviews)
  - is.retracted (keep only those that have NOT been retracted)
 
 Step 2) verify those worked by looking at the number of distinct values
 
 Step 3) filter out additional observations by:
  - publication year (keep only pre-2026)
  - title (remove those with blank titles)
  - primary topic (remove NA) [see code in "02b" for exploration of this variable]
  - fwci (remove NA) [see code in "02b" for exploration of this variable]
  - references_counted (remove any 0 values) [see code in "02b" for exploration of this variable]

Step 4) select out several variables
  - (In Liza's code, she only used work_id, publication year, language, type, and referenced_works_count, but I will keep other variables.)


```{r Steps 1-4: subsetting the "works" rds}

# originally had this filtered to just "article" but based on Wood et al. (2025), decided to also keep "review", "book" and "book-chapter" 

# UPDATE 5/14/2026
    # need to create a "works_filt" to use as a cross-over with the refs
        # comment out the NAIS-themed filter and keep it as "works_filt"

    # and ALSO create one that limits it to NAIS-themed works
        # add the NAIS-themed filter back in and specify file as "works_filt_indig"

works_filt_indig <- works %>% 
  filter(language == "en") %>% 
  filter(type == "article" | type == "review") %>%  # | type == "book" | type == "book-chapter") 
  filter(is_retracted == "FALSE") %>% 
  filter(publication_year != 2026) %>% 
  filter(title != "") %>% 
  filter(!is.na(primary_topic)) %>% 
  filter(!is.na(fwci)) %>% 
  filter(referenced_works_count != 0) %>% 
  filter(str_detect(title, "Indigenous|indigenous|Tribal|tribal|Tribe|tribe|Native") | 
         str_detect(primary_topic, "Indigenous|indigenous|Tribal|tribal|Tribe|tribe|Native")) %>%
  select(-pmid, -mag, -publication_date, -language, -is_retracted, -is_paratext, -biblio_volume, -biblio_issue, -biblio_first_page, -biblio_last_page, -primary_source_type, -primary_source_issn_l, -created_date, -updated_date)


#  want to test to make sure the filtering worked

n_distinct(works_filt$language)  # [1] 1  IT WORKED - becomes [1] 0 with subsequent filtering
n_distinct(works_filt$type)  # [1] 2  IT WORKED - becomes [1] 2 with subsequent filtering
n_distinct(works_filt$is_retracted)  # [1] 1  IT WORKED - becomes [1] 0 with subsequent filtering
n_distinct(works_filt$is_paratext)  # [1] 1  Didn't filter by this variable, but expected this result
                                          # and as expected - - becomes [1] 0 with subsequent filtering

# now I can go back and select out more columns
# end result for now is 21 variables
    # could probably select out more, but want to look at more bibliometric analyses first

saveRDS(works_filt, "data/works_filtered.rds")
saveRDS(works_filt_indig, "data/works_filtered_indig.rds")

```


Moving on to the "refs" rds

Step 5) Remove any self-references and loops 

Step 6) keep only those articles that were cited at least 5 times (per Wood et al. 2025) 

Step 7) Once I have filtered it down to only those references with 5 or more citations, generate a list of just the unique values from both the "citing" and "cited" columns (as in, it's unique to BOTH columns). This step is NEW - before, I was limiting the refs to just those IDs that appeared as both citing and cited, but that would selectively miss older works that are part of the conversation (older articles CAN'T cite newer ones).


```{r Step 5) subsetting the "refs" rds - removing self-refs and loops}

nrow(refs)  # [1] 634527

# remove self references

refs_noselfs <- refs[citing_id != cited_id]
nrow(refs_noselfs)  # [1] 634058 - so, it removed 469 rows

# remove loops - per Liza's code:
  # these shouldn't really be possible
  # unless published in the same issue? And just a few more
  # e.g. A -> B and B -> A

el <- refs_noselfs %>% 
  select(from = citing_id, to = cited_id)

# Liza used a for-loop, but it is taking forever (gave up at 50 minutes)
# Elise's recommendation:
  # use the "which" statement from Liza's for-loop to create an "index"
  # el2 <- el2[!index,] 
  # filter out those rows all at once

# length(unique(el$from))  # [1] 81366
# length(unique(el$to))  # [1] 75009

loop_index <- which(paste(el$from, el$to) %in% paste(el$to, el$from))

refs_clean <- el[!loop_index,]

# checks to see if this approach worked:
  # length(loop_index)  # [1] 5636
  # nrow(refs_reduced)  # [1] 628422
  # nrow(el) - length(loop_index)  # [1] 628422 - IT WORKED!!

length(unique(refs_clean$from))  # [1] 80883
length(unique(refs_clean$to))  # [1] 74746


saveRDS(refs_clean, "data/refs_clean.rds")

```

```{r Steps 6-7) reducing the refs by number of occurrences and generating list of uniques }

refs_citing_only_v2 <- refs_clean %>% 
  count(from, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, citation = from)

n_distinct(refs_citing_only_v2)  # [1] 43025
  
refs_cited_only_v2 <- refs_clean %>% 
  count(to, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, citation = to)

n_distinct(refs_cited_only_v2)  # [1] 32644

nrow(refs_cited_only_v2) + nrow(refs_citing_only_v2)  # [1] 75669

refs_unique <- bind_rows(refs_citing_only_v2, refs_cited_only_v2) %>% 
  unique()

n_distinct(refs_unique)  # [1] 56200

saveRDS(refs_unique, "data/refs_unique.rds")

```


Step 8) Limit the "works_filt" to just those unique references in "refs"

(This is to clean up the metadata/nodelist that will be attached to the network that is built using the refs as the edgelist.  If the "works" are not found in the edgelist, don't need them in the nodelist.)


```{r subsetting the filtered works to match the unique refs}

# UPDATE 5/14/2026
    # need to create a "works_unique" to use as a cross-over with the refs
        # comment out the NAIS-themed filter and keep it as "works_unique"

    # and ALSO create one that limits it to NAIS-themed works
        # add the NAIS-themed filter back in and specify file as "works_unique_indig"

works_unique <- works_filt %>% 
  filter(work_id %in% refs_unique$citation)

nrow(works_unique) # [1] 54575
nrow(works_filt) - nrow(works_unique)  # [1] 48500

print(count(works_unique, is.na(work_id)))  # FALSE 54575, so no "NA" values

## ## ## ## 

works_unique_indig <- works_filt_indig %>% 
  filter(work_id %in% refs_unique$citation)

nrow(works_unique_indig) # [1] 473
nrow(works_filt) - nrow(works_unique_indig)  # [1] 102602
nrow(works_filt_indig) - nrow(works_unique_indig)  # [1] 2481

print(count(works_unique_indig, is.na(work_id)))  # 1:          FALSE 473, no so "NA" values

# so, there are 2481 references in the "works" rds that 
# DO contain the NAIS-relevant keywords, 
      # BUT will NOT be included in the network analysis

# I generated a csv of the 2954 works with NAIS keywords in the topic or title field
    # in the "02b" workflow code

# But I want a separate one of the just the works that will be used for the network analysis

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_indig_unique <- "Indigenous_unique_20260514.csv"
full_output_path_fields <- file.path(output_directory, file_name_indig_unique)

write_excel_csv(works_unique_indig, file = full_output_path_fields)

```

```{r clean up the Environment}

rm(refs_cited_only_v2, refs_citing_only_v2, refs_noselfs)

```


Step 9) With that set of NAIS-themed works, need to create the edge list.

Will use "or" to get the broader landscape of conversation.  (If I use &, then I'm only going to get observations for which BOTH the from and to are part of the NAIS-themed works - meaning, I won't know if there are connections to other fields or not.)

```{r 9) Creating the "indigenous-themed" edgelist}

# will limit the refs (clean) to just those works with NAIS-specific titles or topics

# this will find references that are NOT NAIS-themed but 
    # do reference or are referenced by one of the NAIS-themed works

# MUST intersect the refs_clean with the UNIQUE works
    # if you just use the "filtered" works,
    # you'll get works that DO show up in the refs, 
    # BUT were not referenced >= 5x

refs_or <- refs_clean %>% 
  filter((from %in% works_unique_indig$work_id) | (to %in% works_unique_indig$work_id))  # [1] 4603


# this is a more insular approach - will include
    # only NAIS-themed works that reference or are referenced by other NAIS-themed works

refs_and <- refs_clean %>% 
  filter((from %in% works_unique_indig$work_id) & (to %in% works_unique_indig$work_id))  # [1] 727

nrow(refs_or) - nrow(refs_and)   
  # [1] 3876 - the number of works for which either the "from" or "to" 
                  # do not ALSO use NAIS-keywords in the title or "primary_topic" field

saveRDS(refs_or, "data/edgelist_indig_full.rds")
saveRDS(refs_and, "data/edgelist_indig_insular.rds")


```


Step 10) Need to a) repeat the process of getting only unique values of the "refs_or" subset, then b) filter the "works" to just those IDs to generate the nodelist.  



```{r 10a) getting just the refs with an NAIS-coded title or primary topic}

# get the unique list of refs in the "refs_or" object

refs_citing_only_or <- refs_or %>% 
  count(from, sort = TRUE, name = "tally") %>% 
  select(-tally, citation = from)

n_distinct(refs_citing_only_or)  # [1] 1793
  
refs_cited_only_or <- refs_or %>% 
  count(to, sort = TRUE, name = "tally") %>% 
  select(-tally, citation = to)

n_distinct(refs_cited_only_or)  # [1] 1941

nrow(refs_cited_only_or) + nrow(refs_citing_only_or)  # [1] 3734

refs_unique_or <- bind_rows(refs_citing_only_or, refs_cited_only_or) %>% 
  unique()

n_distinct(refs_unique_or)  # [1] 3250

# saveRDS(refs_unique_or, "data/refs_unique_Indig.rds")


```


```{r 10b) subsetting the filtered works to match the unique refs}

works_indig_or <- works_filt %>% 
  filter(work_id %in% refs_unique_or$citation) 

saveRDS(works_indig_or, "data/nodelist_Indig_full.rds")

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_indig_or <- "Nodelist_Indig_full_20260514.csv"
full_output_path_fields <- file.path(output_directory, file_name_indig_or)

write_excel_csv(works_indig_or, file = full_output_path_fields)

```
Step 10) As part of considering whether or not to filter based on the different "primary_topic*" variables, I also wanted to assess the distribution of "Indigenous" specific language in both the title and primary topic.  Takeaway:  In the filtered works that appear in the unique list of references, there are 473 observations that have "Indigenous|indigenous|Tribal|tribal|Tribe|tribe|Native") in either the title or primary topic field - which equates to 0.88% of the works.  (In the full list of filtered references, including those not found in the unique "refs", there are 2954 observations - which equates to 2.8%.)  [see code in "02b" for exploration of this occurrence] 




```{r code for writing csvs}

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```


