---
title: " 02b_bibliometrics_data_subsetting_exploration"
author: "Mackenzie Johnson"
date: "20260401"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
library(tidyverse)
library(stringr)
```


```{r code for writing csvs}

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```



```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

```

The data subsetting step of this project was an iterative process.  After initially subsetting based on language, type, retraction, and publication year, and removing those with blank values for the title, I wanted to explore several other attributes of the data.  

1) NA values for either the primary topic or fwci - I then filtered the "filt_works" object based on this exploration 

```{r Data Exploration - "filt_works" - NA removal} 

print(count(filt_works, is.na(title)))  # 1:        FALSE 142103 -- as in, all of them, so no NA values

print(count(filt_works, is.na(fwci)))  # 2:        TRUE  10694
print(count(filt_works, is.na(primary_topic)))  # 2: TRUE   10680  
print(count(filt_works, is.na(primary_topic_subfield)))  # 2: TRUE   10680
print(count(filt_works, is.na(primary_topic_field)))  # 2: TRUE   10680
print(count(filt_works, is.na(primary_topic_domain)))  # 2: TRUE   10680

#  after filtering out the NA values for "primary_topic" - all the "primary_topic_..." variables have no NA values
# and "fwci" only has 14 values, so I'll filter those out 
# can see from Global Environments it worked - now there are 131409 observations

```


2) Works with 0 values for "referenced_works_count" - I then filtered the "filt_works" object based on this exploration

```{r Data Exploration - "filt_works" - works with no references }

print(min(filt_works$referenced_works_count))
print(count(filt_works, referenced_works_count == 0))

    # referenced_works_count == 0      n
                            # <lgcl>  <int>
    # 1:                       FALSE 103075
    # 2:                        TRUE  28334  

count_no_refs <- as.integer(count(filt_works[referenced_works_count == 0]))  

tally_no_refs_by_source_group <- filt_works %>% 
  filter(referenced_works_count == 0) %>% 
  group_by(source_group) %>%  # group_by(source_id) -- gets the same tally as "by_journal" but now with source_id, not journal name
  summarise(n = n()) %>% 
  mutate(pct = n/count_no_refs) %>% 
  print()


tally_no_refs_by_journal <- filt_works %>% 
  filter(referenced_works_count == 0) %>% 
  group_by(primary_source_display_name) %>% 
  summarise(n = n()) %>% 
  mutate(pct = n/count_no_refs) %>% 
  arrange(desc(n))
  

count(referenced_works_count, sort = FALSE, name = "tally")

length(filt_works$referenced_works_count[filt_works$referenced_works_count < 10])  # 43782
length(filt_works$referenced_works_count[filt_works$referenced_works_count >= 10])  # 87627

no_refs_in_works <- filt_works %>% 
  filter(referenced_works_count == 0)

str(no_refs_in_works)  # Classes ‘data.table’ and 'data.frame':	25529 obs. of  21 variables

no_refs_overlap <- no_refs_in_works %>% 
  filter(work_id %in% refs_filtered_overlap$citation)

str(no_refs_overlap)  # Classes ‘data.table’ and 'data.frame':	0 obs. of  21 variables

no_refs_in_works_open <- filt_works %>% 
  filter(referenced_works_count == 0) %>% 
  filter(oa_status != "closed")

str(no_refs_in_works_open)  # Classes ‘data.table’ and 'data.frame':	2422 obs. of  21 variables
length(no_refs_in_works_open$has_fulltext[no_refs_in_works_open$has_fulltext == "TRUE"])  # 720 - these have full text
length(no_refs_in_works_open$oa_is_oa[no_refs_in_works_open$oa_is_oa == "TRUE"]) # [1] 2422 - good, SHOULD match the length of the df
n_distinct(no_refs_in_works_open$oa_is_oa)  # [1] 1 - yep, apparently they are all OA - that 1 value is "TRUE")


# OK, so seems like none of the references with zero values for the "referenced_works_count" show up in the "refs" rds
# even if their OA status is NOT closed


one_refs_in_works <- filtered_works %>% 
  filter(referenced_works_count == 1)

str(one_refs_in_works)  # Classes ‘data.table’ and 'data.frame':	2358 obs. of  21 variables

one_refs_overlap <- one_refs_in_works %>% 
  filter(work_id %in% refs$cited_id)

str(one_refs_overlap)  # Classes ‘data.table’ and 'data.frame':	1005 obs. of  21 variables

# BUT, as soon as there is a single referenced work, then there is overlap

```


3) Verification that intersecting the filtered works with the overlapping references worked.  

This step used %in% to select just those observations in the the "filt_works" object that are found in the "refs_filt_overlap" object.  But the number of observations of the new object did NOT match the number in the list of overlapping references; the following code shows how the intersection was verified.

(Prior to using the %in% approach for this step, I tested the different types of joins to see if either of them seemed like a better option.  That code is in the "02_bibliometrics_data_subsetting_tests" rmd.)


```{r need to find the 57 citations from the overlapping references that aren't in the filtered works}

refs_to_join <- refs_filt_overlap %>% 
  select(work_id = citation_id)

joined_refs_works <- full_join(filt_works, refs_to_join, by = "work_id")

nrow(joined_refs_works)  # [1] 103132
nrow(joined_refs_works) - nrow(filt_works)  # [1] 57

print(count(joined_refs_works, is.na(work_id)))  # 1:  FALSE 103132 (so no NA values in the work_id)
print(count(joined_refs_works, is.na(source_id)))   # 2: TRUE     57

missing_refs_only <- joined_refs_works %>%
  filter(is.na(source_id))

# I want to explore these manually since there are only 57 of them
# need to write a csv and do this in Excel

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name <- "Refs_not_in_Works.csv"
full_output_path <- file.path(output_directory, file_name)


##  ACTUALLY -- there's a faster way to do this...

refs_not_in_works <- works %>%
  filter(work_id %in% missing_refs_only$work_id)

write_excel_csv(refs_not_in_works, file = full_output_path)

# OK, found them!  
# in the spreadsheet, I highlighted why each would have been filtered out
# so, should definitely use either the %in% or one of the joins that does NOT keep the "refs" that aren't in the "works"

```



4) I considered filtering out rows based on primary topic information

I did this both before and after I intersected the "filt_works" with the "refs_filt_overlap".  The results were either copied and pasted into Excel or exported as a csv and copied into a spreadsheet ("Bilbiometrics_R_to_Excel").  


```{r investigating the "primary topic" variables of the filtered works}

# tally_primary_topic_field_filtered <- count(filt_works, primary_topic_field, sort = TRUE, name = "tally")
# str(tally_primary_topic_field_filtered)  # Classes ‘data.table’ and 'data.frame':	26 obs. of  2 variables

n_distinct(filt_works$primary_topic)  # [1] 2075
n_distinct(filt_works$primary_topic_domain)  # [1] 4
n_distinct(filt_works$primary_topic_field)  # [1] 26
n_distinct(filt_works$primary_topic_subfield)  # [1] 199

n_distinct(filt_works$primary_topic_subfield[filt_works$primary_topic_field == "Social Sciences"])  # [1] 21 
n_distinct(filt_works$primary_topic[filt_works$primary_topic_field == "Social Sciences"])  # [1] 535

tally_by_primary_field <- filt_works %>% 
  group_by(primary_topic_field) %>%
  # mutate(n = summarise(n(primary_topic))) %>%  # This does NOT work
  mutate(n_subfields = n_distinct(primary_topic_subfield)) %>% 
  mutate(n_topics = n_distinct(primary_topic)) %>% 
  select(primary_topic_field, n_subfields, n_topics) %>% 
  distinct(primary_topic_field, .keep_all = TRUE) # %>% 
  # arrange(desc(n))

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(tally_by_primary_field, file = full_output_path_fields)


# might be helpful to look at the works just for some of the fields, 
# or at least to look at the different subfield frequencies


prim_field_biochem <- filt_works %>% 
  filter(primary_topic_field == "Biochemistry, Genetics and Molecular Biology")

prim_field_socsci <- filt_works %>% 
  filter(primary_topic_field == "Social Sciences")

prim_field_envsci <- filt_works %>% 
  filter(primary_topic_field == "Environmental Science")

prim_field_econ <- filt_works %>% 
  filter(primary_topic_field == "Economics, Econometrics and Finance")

prim_field_agbio <- filt_works %>% 
  filter(primary_topic_field == "Agricultural and Biological Sciences")


tally_biochem <- prim_field_biochem %>% 
  count(source_group, sort = TRUE, name = "tally") %>% 
  mutate(pct = tally / nrow(prim_field_biochem))

biochem_topics <- prim_field_biochem %>% 
  group_by(primary_topic_subfield) %>% 
  count(primary_topic, sort = TRUE, name = "n_topics") %>% 
  mutate(n_subfields = count(primary_topic_subfield, ))

```

```{r investigating the "primary topic" variables of the filtered AND overlapping works}

n_distinct(filt_overlap_works$primary_topic)  # [1] 577
n_distinct(filt_overlap_works$primary_topic_domain)  # [1] 4
n_distinct(filt_overlap_works$primary_topic_field)  # [1] 19
n_distinct(filt_overlap_works$primary_topic_subfield)  # [1] 105

n_distinct(filt_overlap_works$primary_topic_subfield[filt_overlap_works$primary_topic_field == "Social Sciences"])  # [1] 17 
n_distinct(filt_overlap_works$primary_topic[filt_overlap_works$primary_topic_field == "Social Sciences"])  # [1] 183

overlap_by_primary_field <- filt_overlap_works %>% 
  group_by(primary_topic_field) %>%
  summarise(n = n())
  # mutate(n = summarise(n(primary_topic))) %>%  # This does NOT work
  mutate(n_subfields = n_distinct(primary_topic_subfield)) %>% 
  mutate(n_topics = n_distinct(primary_topic)) %>% 
  select(primary_topic_field, n_subfields, n_topics) %>% 
  distinct(primary_topic_field, .keep_all = TRUE) # %>% 
  # arrange(desc(n))
  

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics_overlap.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(overlap_by_primary_field, file = full_output_path_fields)

# --- Deeper dive into top 8 fields --- #

prim_field_socsci_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Social Sciences")

prim_field_econ_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Economics, Econometrics and Finance")

prim_field_envirosci_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Environmental Science")

prim_field_energy_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Energy")

prim_field_eng_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Engineering")

prim_field_busmgmt_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Business, Management and Accounting")

prim_field_decsci_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Decision Sciences")

prim_field_health_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Health Professions")

prim_field_agbio_overlap <- filt_overlap_works %>% 
  filter(primary_topic_field == "Agricultural and Biological Sciences")

# --- Base R option --- #
output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name <- "SocSci_overlap_subfields2.csv"
full_output_path_fields <- file.path(output_directory, file_name)

write_excel_csv(as.data.frame(table(prim_field_socsci_overlap$primary_topic_subfield)), file = full_output_path_fields)

# --- #

# --- Tidy option --- #

# run for each of the 9 fields above
# no need to copy code, will just replace the necessary components

agbio_overlap_topics <- prim_field_agbio_overlap %>% 
  group_by(primary_topic_subfield) %>% 
  summarise(n_subfield = n())
 

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name <- "AgBio_overlap_subfields.csv"
full_output_path_fields <- file.path(output_directory, file_name)

write_excel_csv(agbio_overlap_topics, file = full_output_path_fields)

```


```{r tallying "Indigenous"}

# run 1:  BEFORE limiting the "filt_works" to just those that appear in the overlapping reference list
  ## each command starts by piping in "filt_works" 

# run 2:  AFTER limiting the "filt_works" to just those that appear in the overlapping reference list
  ## each command starts by piping in "filt_overlap_works" 

indig_title_topic_or <- filt_overlap_works %>% 
  filter(str_detect(title, "Indigenous|indigenous|tribal|tribe|Tribe|Native") | 
           str_detect(primary_topic, "Indigenous|indigenous|tribal|tribe|Tribe|Native")) %>%  # space between terms OK!
  group_by(primary_topic_field) %>%  # group_by(primary_topic_field)
  summarise(n_Indig = n())

# for a deeper look at the works that appear in both the citing and cited columns of the original "refs" 
# AND also have "Indigenous" or something along those lines in either the title or the primary_topic field:
# use the code above, but only before the "group_by"

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_indig <- "Overlap_Indigenous.csv"
full_output_path_fields <- file.path(output_directory, file_name_indig)

write_excel_csv(indig_title_topic_or, file = full_output_path_fields)



# ---------------
# just need the above code
# verified that the above captures everything correctly using the intermediaries below
# did so for both runs

indig_prim_topic <- filt_overlap_works %>% 
  filter(str_detect(primary_topic, "Indigenous|indigenous|tribal|tribe|Tribe|Native")) %>%  # no spaces between terms!
  group_by(primary_topic_field) %>%  # group_by(primary_topic_field)
  summarise(n_Indig = n())

indig_title <- filt_overlap_works %>% 
  filter(str_detect(title, "Indigenous|indigenous|tribal|tribe|Tribe|Native")) %>% 
  group_by(primary_topic_field) %>%  # group_by(primary_topic_field)
  summarise(n_Indig = n())


indig_title_topic_both <- filt_overlap_works %>% 
  # group_by(primary_topic_field) %>%
  filter(str_detect(title, "Indigenous|indigenous|tribal|tribe|Tribe|Native") & 
           str_detect(primary_topic, "Indigenous|indigenous|tribal|tribe|Tribe|Native")) %>% 
  group_by(primary_topic_field) %>%  # group_by(primary_topic_field)
  summarise(n_Indig = n())

```


```{r Elise's help}

# Elise reminded me of a simpler way to get the counts - use "table"

  # but to write the table to csv, need to make it a dataframe
  # as.data.frame(table(prim_field_biochem$primary_topic_subfield))

# just need to write them to a csv

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name <- "AgBio_subfields.csv"
full_output_path_fields <- file.path(output_directory, file_name)

write_excel_csv(as.data.frame(table(prim_field_agbio$primary_topic_subfield)), file = full_output_path_fields)

table(filt_works$primary_topic)
table(prim_field_biochem$primary_topic)
table(prim_field_biochem$primary_topic_subfield)

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```

```{r Elise's help} 

# table(filt_works$)

```





```{r subsetting the "refs" rds using %in% }

refs_citing_only_2 <- refs %>% 
  count(citing_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, from = citing_id)
  
refs_cited_only_2 <- refs %>% 
  count(cited_id, sort = TRUE, name = "tally") %>% 
  filter(tally >= 5) %>% 
  select(-tally, to = cited_id)


citing_in_cited <- refs_citing_only_2 %>% 
  filter(from %in% refs_cited_only_2$to)

cited_in_citing <- refs_cited_only_2 %>% 
  filter(to %in% refs_citing_only_2$from)


# now that I have isolated both the "citing_id" works and the "cited_id" works into their own dataframes, 
# filtered them to just those works that appear 5 or more times
# and have renamed them with the same column name
# I should be able to intersect them

# refs_filtered_overlap_2 <- intersect(refs_citing_only_2, refs_cited_only_2)  # doesn't work b/c columns don't have same name

typeof(filtered_refs_overlap) # [1] "list"
class(filtered_refs_overlap)  # [1] "data.table" "data.frame"

n_distinct(refs_filtered_overlap$citation)  # [1] 18977 - this matches the length of the dataframe!

# um - I think it worked!  
# given that the number of distinct values equals the number of observations
# and given the below
    # tally_filtered_refs_overlap <- count(filtered_refs_overlap, citation, sort = TRUE, name = "tally")
    # min(tally_filtered_refs_overlap$tally)  # [1] 1
    # max(tally_filtered_refs_overlap$tally)  # [1] 1

```



