---
title: "03_bibliometrics_networks"
author: "Mackenzie Johnson"
date: "2026-04-09"
output: html_document
---

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
# knitr::opts_chunk$set(include = FALSE)

```

```{r libraries}

# install.packages("pbapply")
# install.packages("slam")
# install.packages("ggraph")

library(data.table)
library(ggplot2)
library(ggraph)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)
library(stringr)
library(tidyverse)

```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# raw data

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

# edgelist

el_big <- readRDS(file = "data/edgelist_soc_env_sci_or")
el_sm <- readRDS(file = "data/edgelist_soc_env_sci_and")

```

[UPDATE:  4/27/2026

Started trying to make sense of the network again now that I have a new edgelist.  I brought the clustering algorithm out of the for-loop.  Made it through 2c.]


Step 1)  Create a citation graph following Wood et al. (2025) - code is available in Liza Wood's Github repository "tamar"
        a) start with a full graph
        b) simplify it by removing any multiple edges or loop edges - BUT!  WARNING!  Elise says "simplify" is notoriously wonky

```{r 1) create citation graph}

# this block is essentially copy-pasted from Liza's code
  # only difference is I added some annotations

cit_graph <- graph_from_edgelist(as.matrix(el_big), directed = T)

# str(cit_graph)

# Class 'igraph'  hidden list of 10
#  $ : num 94232

# result is a "Large igraph (69398 elements, 11.1 MB)

cit_graph <- igraph::simplify(cit_graph, remove.multiple = T, remove.loops = T)
  # remove.multiple	Logical, whether the multiple edges are to be removed
  # remove.loops	Logical, whether the loop edges are to be removed

# result is a "Large igraph (94232 elements, 5 MB)

```


Step 2) Identify clusters
        a) Find connected components of the citation graph
        b) For those components from the main cluster (with 69,041 elements out of 69,144), create subgraph
        c) run clustering algorithm on the main cluster - start with Louvain
        d) examine modularity
        e) examine Page Rank estimates of citation importance
    

```{r 2a) find connected components}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus"
  # added some annotations

clus <- components(cit_graph)  # List of 3: $ membership, $ csize, $ no

# result is a "Large list (3 elements, 5.6 MB)

str(clus)
  # List of 3
  # $ membership: Named num [1:69398] 1 1 1 1 ...
    # ..- attr(*, "names")= [1:69398] "W1579625741"  ...
  # $ csize     : num [1:21] 69191 16 6 15 6 ...
  # $ no        : num 21

# --- grouping/summarizing doesn't work --- "

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(clus$membership)

#       1     2     3     4     5     6     7     8     9     
#   69191    16     6    15     6    40    14    10     7    

# 10    11    12    13    14    15  16    17    18    19    20    21 
#  6     6    10     6     9     9   7    11     7     8     6     8

# translation:  there are 21 clusters - 69191 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_component <- names(clus$membership)[clus$membership == 1]

str(connected_component)  #  [1:69191]  
    # --> which means 207 works were NOT INCLUDED in the main cluster (69398-69191)

unconnecteds <- names(clus$membership)[clus$membership != 1]  #  chr [1:1729]
    # can eventually use this list to find the "outliers"
    # as Liza did


ggraph(cit_graph, layout = "fr") +
geom_edge_fan(
end_cap = circle(1, "mm"),
color = "#000000",
width = 0.3,
arrow = arrow(angle = 15, length = unit(0.07, "inches"), ends = "last", type = "closed"))+
geom_node_point(alpha = 0.2) +
labs(title = "Citation Graph: Limited to Social & Environmental Sciences") + theme_void()

```

```{r 2b) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus" but no other changes

if(any(clus$membership>1)){
  cit_graph_main <- induced_subgraph(cit_graph,v = which(V(cit_graph)$name %in% connected_component))
  cit_graph <- cit_graph_main
}

cit_graph_ud <- as.undirected(cit_graph, 'collapse')

```


```{r 2c) run clustering algorithm}

# this block is essentially copy-pasted from Liza's code

# recall that she had a much higher seed value under certain call conditions
      # if(LOCATION == 'data/query_EG/' & ADD_SUPPL_WOS_REFS == F){
        #  pref_seed <- 4812
        #  } else{

# will try running it with the set.seed set to 12 
    # idea is to seed a pseudo random number generator using a reproducible seed 
    # select 1000 numbers with equal weighting from the set 0:9999
    # why do we need this?

set.seed(12)  # Liza comment: # and 16 got to 8, dropped PES and joined it with PE
seeds <- sample.int(9999, 1000)

mod <- list()
for(i in 1:length(seeds)){
  set.seed(seeds[i])
      # why is the clustering algo in the for-loop? 
      # Not affected by iteration, move to outside of loop for efficiency
  louv_soc_env <- cluster_louvain(cit_graph_ud)
  mod$mod[i] <- modularity(louv_soc_env)
  mod$min_n[i] <- min(table(louv_soc_env$membership))
}
mods <- unlist(mod$mod)
min_ns <- unlist(mod$min_n)

saveRDS(louv_soc_env, "Louvain_soc_env")
# louv <- readRDS(file = "data/Louvain_full")
    
head(mods)  # [1] 0.6924431 0.6934947 0.6948436 0.6965930 0.6892218 0.6947261
tail(mods)  # [1] 0.6943069 0.6962361 0.6948948 0.6928662 0.6887062 0.6983792

head(louv_soc_env$modularity)  # [1] 0.6204364 0.6940245 0.6983299 0.6983792
tail(louv_soc_env$modularity)  # [1] 0.6204364 0.6940245 0.6983299 0.6983792
# print(louv_soc_env$modularity)

```

```{r 2d) examine modularity and degrees}

# this block is essentially copy-pasted from Liza's code

pref_seed <- seeds[which(mods == max(mods))]

set.seed(pref_seed)  # Liza comment:  # 4262 with min size of 80, 2925 with min size of 30
# louv <- readRDS(file = "data/Louvain_soc_env")
V(cit_graph_main)$louv_cluster <- louv_soc_env$membership

# modularity measure
    # B_ij = A_ij - γ k_i k_j / (2m)  == for UNDIRECTED graphs
    # B_ij = A_ij - γ k^out_i k^in_j / m == for DIRECTED graphs

modularity(louv_soc_env)
B = modularity_matrix(cit_graph_main)

    # Error in modularity_matrix(cit_graph_main) :
        # At vendor/cigraph/src/core/vector.pmt:475 : Cannot reserve space for vector. Out of memory

# out-degree, in-degree, total-degree 

summary(degree(cit_graph_main,mode = 'out'))
  # Min.    1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 0.000   1.000     4.000   6.782 9.000   190.000 

summary(degree(cit_graph_main,mode = 'in'))
  #  Min.  1st Qu.   Median     Mean    3rd Qu.     Max. 
  #  0.000  1.000    3.000    6.782     7.000    1068.000

summary(degree(cit_graph_main, mode = "all"))
  #  Min. 1st Qu.  Median    Mean   3rd Qu.    Max. 
  # 1.00    3.00    8.00    13.56   17.00     1072.00

# Add network summary
table(louv_soc_env$membership)
mean(table(louv_soc_env$membership))  # [1] 1976.886
median(table(louv_soc_env$membership))  # [1] 107
min(table(louv_soc_env$membership))  # [1] 5
max(table(louv_soc_env$membership))  # [1] 12561

```



```{r e) PageRank measure of citation importance}

prank <- page_rank(cit_graph)
V(cit_graph)$prank <- unlist(as.vector(prank$vector))

betw <- betweenness(cit_graph)
V(cit_graph)$betweenness <- as.vector(betw)
betw_group <- data.table(node = V(cit_graph)$name,
                         between = betw,
                         group = V(cit_graph)$louv_cluster)
betw_group[,grank:=rank(-between,ties.method = 'average'),by=.(group)]


V(cit_graph)$high_between <- V(cit_graph)$name %in% betw_group[grank<=4,]$node

#V(cit_graph)[betw == max(betw)]
prank_max <- data.table(node = V(cit_graph)$name,group = unlist(V(cit_graph)$louv_cluster),pagerank = unlist(V(cit_graph)$prank))
prank_max_by_group <- prank_max[,max(pagerank),by=.(group)]

max_by_group = dplyr::left_join(prank_max,prank_max_by_group)[pagerank==V1,]

V(cit_graph)$highest_in_group <- V(cit_graph)$name %in% max_by_group$node

saveRDS(cit_graph, paste0(path_location,'cit_graph.rds'))

```

