---
title: "03_bibliometrics_networks"
author: "Mackenzie Johnson"
date: "2026-04-09"
output: html_document
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
# install.packages("pbapply")
# install.packages("slam")

library(data.table)
library(ggplot2)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)
library(stringr)
library(tidyverse)

```


```{r code for writing csvs}

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```



```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData1}

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

# edgelist
el <- readRDS(file = "data/network_edgelist_reduced")


```

Step 1)  Create a citation graph following Wood et al. (2025) - code is available in Liza Wood's Github repository "tamar"
        a) start with a full graph
        b) simplify it 

```{r create citation graph}

cit_graph <- graph_from_edgelist(as.matrix(el), directed = T)

# str(cit_graph)

# Class 'igraph'  hidden list of 10
#  $ : num 69144

# result is a "Large igraph (69144 elements, 12.4 MB)

cit_graph <- igraph::simplify(cit_graph, remove.multiple = T, remove.loops = T)

# result is a "Large igraph (69144 elements, 5 MB)

```


Step 2) Identify clusters
        a) Find connected components of the graph
        b) For those components with a membership size larger than 1 (i.e., connected to at least one other node), create a subgraph
        c) run clustering algorithm
        d) examine modularity
    

```{r a) find connected components}

clus <- components(cit_graph)

# result is a "Large list (3 elements, 5.5 MB)

# str(clus)
  # List of 3
  # $ membership: Named num [1:69144]
    # ..- attr(*, "names")= chr [1:69144] "W2599301038" ...
  # $ csize     : num [1:9]
  # $ no        : num 9

# clus_summary <- clus %>% 
#   group_by(membership) %>% 
#  summarise(n = n())

# Error in UseMethod("group_by") : 
  # no applicable method for 'group_by' applied to an object of class "list"

table(clus$membership)

#     1     2     3     4     5     6     7     8     9 
# 69041     9    11    19    15     9    15    12    13 

connected_component <- names(clus$membership)[clus$membership!=1]
# str(connected_component)  #  chr [1:103]

```

```{r b) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code

if(any(clus$membership>1)){
  cit_graph_main <- induced_subgraph(cit_graph,v = which(V(cit_graph)$name %in% connected_component))
  cit_graph <- cit_graph_main
}

cit_graph_ud <- as.undirected(cit_graph, 'collapse')
```

```{r c) run clustering algorithm}

# this block is essentially copy-pasted from Liza's code

set.seed(12) 
    seeds <- sample.int(9999, 1000)

 mod <- list()
    for(i in 1:length(seeds)){
      set.seed(seeds[i])
      louv <- cluster_louvain(cit_graph_ud)
      mod$mod[i] <- modularity(louv)
      mod$min_n[i] <- min(table(louv$membership))
    }
 
 mods <- unlist(mod$mod)
    min_ns <- unlist(mod$min_n)

```

```{r d) examine modularity}
pref_seed <- seeds[which(mods == max(mods))]

set.seed(pref_seed) # 4262 with min size of 80, 2925 with min size of 30
louv <- cluster_louvain(cit_graph_ud)
V(cit_graph)$louv_cluster <- louv$membership

# when I try to view "louv" - get an error message:
# Error in names[[i]] : subscript out of bounds

# modularity measure
modularity(louv)
B = modularity_matrix(cit_graph)

summary(degree(cit_graph,mode = 'out'))
  # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 0.00    0.00    1.00    1.35    1.00   13.00

summary(degree(cit_graph,mode = 'in'))
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  #  0.00    0.00    1.00    1.35    1.00   12.00

summary(degree(cit_graph))
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 1.000   1.000   1.000   2.699   1.000  24.000

# Add network summary
table(louv$membership)
mean(table(louv$membership))
median(table(louv$membership))
min(table(louv$membership))  # [1] 9
max(table(louv$membership))  # [1] 19

```

```{r d) cluster descriptions}

prank <- page_rank(cit_graph)
V(cit_graph)$prank <- unlist(as.vector(prank$vector))

betw <- betweenness(cit_graph)
V(cit_graph)$betweenness <- as.vector(betw)
betw_group <- data.table(node = V(cit_graph)$name,between = betw,group = V(cit_graph)$louv_cluster)
betw_group[,grank:=rank(-between,ties.method = 'average'),by=.(group)]


V(cit_graph)$high_between <- V(cit_graph)$name %in% betw_group[grank<=4,]$node

#V(cit_graph)[betw == max(betw)]
prank_max <- data.table(node = V(cit_graph)$name,group = unlist(V(cit_graph)$louv_cluster),pagerank = unlist(V(cit_graph)$prank))
prank_max_by_group <- prank_max[,max(pagerank),by=.(group)]

max_by_group = dplyr::left_join(prank_max,prank_max_by_group)[pagerank==V1,]

V(cit_graph)$highest_in_group <- V(cit_graph)$name %in% max_by_group$node

saveRDS(cit_graph, paste0(path_location,'cit_graph.rds'))

```
