---
title: "03_bibliometrics_networks_indig_full"
author: "Mackenzie Johnson"
date: "2026-04-09"
output: html_document
---

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
# knitr::opts_chunk$set(include = FALSE)

```

```{r libraries}

# install.packages("pbapply")
# install.packages("slam")
# install.packages("ggraph")

library(data.table)
library(ggplot2)
library(ggraph)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)
library(stringr)
library(tidyverse)

```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# edgelist

el_big <- readRDS(file = "data/edgelist_indig_or")  # 4603 observations of 2 variables
# el_sm <- readRDS(file = "data/edgelist_indig_and")  # 727 observations of 2 variables

```

[UPDATE:  4/27/2026 and 5/12/2016

Started trying to make sense of the network again now that I have a new edgelist.  I brought the clustering algorithm out of the for-loop.  Made it through 2c.]


Step 1)  Create a citation graph following Wood et al. (2025) - code is available in Liza Wood's Github repository "tamar"
        a) start with a full graph
        b) simplify it by removing any multiple edges or loop edges - BUT!  WARNING!  Elise says "simplify" is notoriously wonky
        c) test if it's "connected"

        
Regarding "is_connected()"

From "networkx.org" documentation for doing network analysis in Python:
  * "A graph is connected if, for every pair of distinct nodes, there is a path between them. If there is a pair of nodes for which such path does not exist, the graph is not connected (also referred to as “disconnected”)"

From github:  https://olizardo.github.io/networks-textbook/lesson-graphs-connectivity.html
  * "A graph is CONNECTED if there exists a (finite) path between each pair of nodes.  A graph is DISCONNECTED if there does not exist a path connecting every pair of nodes"
  
From github:  https://yunranchen.github.io/intro-net-r/igraph.html
  * "Connected: if there is a path from any vertex to any other.  Unconnected: if not connected. An unconnected graph has multiple components." 
       
       
```{r 1-big) create citation graphs}

# this block is essentially copy-pasted from Liza's code
  # only difference is I added some annotations

# this run is for articles with NAIS-themed titles or primary_topic
    # in the FULL list of works (after filtering/before limiting to the unique refs)
    # meaning: these works might reference works that do NOT have NAIS-themed titles or topics


cit_graph_big <- graph_from_edgelist(as.matrix(el_big), directed = T)  # List of 3250

# str(cit_graph_big)


cit_graph_big <- igraph::simplify(cit_graph_big, remove.multiple = T, remove.loops = T)
  # remove.multiple	Logical, whether the multiple edges are to be removed
  # remove.loops	Logical, whether the loop edges are to be removed

# Liza didn't do this next step, but I want to check

is_connected(cit_graph_big, mode = "weak")  # [1] FALSE

# FALSE means that it is disconnected (meaning, there are gaps between some of the nodes)

count_components(cit_graph_big)  # [1] 25 --- matches component details in the next chunk

# to get a histogram for the maximal connected component sizes: 
  # component_distribution(cit_graph_big)

```


Step 2) Network components ("maximal induced subgraph that is connected")
        a) Find connected components of the citation graph and visualize the network
        b) For those components from the main cluster (with 69,041 elements out of 69,144), create subgraph
        c) run clustering algorithm on the main cluster - start with Louvain
        d) examine modularity
        e) examine Page Rank estimates of citation importance
    

```{r 2a-big) find connected components of the larger graph}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus"
  # added some annotations

# despite the "FALSE" result above for "is_connected" 
# ran the "components" function using both "weak" and "strong" modes
    # the "weak" run yielded the same thing as not specifiying the mode, 
    # the "strong" run yielded 3250 clusters - seems like a Directed Acyclic Graph (?)
 

clus_big <- components(cit_graph_big, mode = "weak")  # List of 3: $ membership, $ csize, $ no
str(clus_big)
  # List of 3
  # $ membership: Named num [1:3250] 1 1 1 1 1 1 1 1 1 1 ...
    # ..- attr(*, "names")= chr [1:3250] "W1579625741" "W1995811201" ...
  # $ csize     : num [1:25] 3013 15 14 10 6 ... ...
  # $ no        : num 25

# --- grouping/summarizing doesn't work --- "

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(clus_big$membership)

# 1       2    3    4    5    6    7    8    9   10   11   12   13   14   15   16   17   18   19 
# 3013   15   14   10    6    7   10    6    7   10   12    9   11   11   12    6    6   17   18 

#   20   21   22   23   24   25 
#   12    9    7    9    7    6 

# translation:  there are 25 clusters - 3013 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_indig_full <- names(clus_big$membership)[clus_big$membership == 1]

str(connected_indig_full)  #  [1:3013]  
    # --> which means 237 works were NOT INCLUDED in the main cluster (3250-3013)

# Got some visualization code from Sara!  

ggraph(cit_graph_big, layout = "fr") +
geom_edge_fan(
end_cap = circle(1, "mm"),
color = "#000000",
width = 0.3,
arrow = arrow(angle = 15, length = unit(0.07, "inches"), ends = "last", type = "closed"))+
geom_node_point(alpha = 0.2) +
labs(title = "NAIS-themed Citations (full works)") + theme_void()

```


Before moving on, I tried figuring out a for-loop to get the cluster membership for each work_id.

It didn't go well...need to come back to this.

unconnecteds_names <- names(clus$membership)[clus$membership != 1]  #  chr [1:1729]
    # can eventually use this list to find the "outliers"
    # as Liza did

clus_names <- names(clus$membership)[clus$membership != 1]
clus_num <- clus$no

cluster_list <- clus$membership



cluster_assign <- for(i in cluster_list) {
  which(clus$membership == i)
}

length(clus$membership)


```{r 2b-big) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus" but no other changes

# the following code is for extracting the main blob, 
# but I also want to see what happens when I run it on the full network
    # especially given the appearance of the edgelist map below

if(any(clus_big$membership>1)){
  cit_graph_big_main <- induced_subgraph(cit_graph_big,v = which(V(cit_graph_big)$name %in% connected_component_big))
  cit_graph_big <- cit_graph_big_main
}

cit_graph_big_ud <- as.undirected(cit_graph_big, 'collapse')

```


```{r 2c-big) run clustering algorithm}

# this block is essentially copy-pasted from Liza's code

# recall that she had a much higher seed value under certain call conditions
      # if(LOCATION == 'data/query_EG/' & ADD_SUPPL_WOS_REFS == F){
        #  pref_seed <- 4812
        #  } else{

# will try running it with the set.seed set to 12 
    # idea is to seed a pseudo random number generator using a reproducible seed 
    # select 1000 numbers with equal weighting from the set 0:9999
    # why do we need this?

set.seed(12)  # Liza comment: # and 16 got to 8, dropped PES and joined it with PE
seeds <- sample.int(9999, 1000)

mod_big <- list()
for(i in 1:length(seeds)){
  set.seed(seeds[i])
      # why is the clustering algo in the for-loop? 
      # Not affected by iteration, move to outside of loop for efficiency
  louv_indig_big <- cluster_louvain(cit_graph_big_ud)
  mod_big$mod[i] <- modularity(louv_indig_big)
  mod_big$min_n[i] <- min(table(louv_indig_big$membership))
}
mods <- unlist(mod_big$mod)
min_ns <- unlist(mod_big$min_n)

# saveRDS(louv_indig, "Louvain_indig")
# louv <- readRDS(file = "data/Louvain_full")
    
head(mods)  # [1] 0.8170753 0.8162952 0.8183118 0.8163955 0.8195041 0.8157186
tail(mods)  # [1] 0.8154065 0.8164846 0.8171450 0.8147561 0.8165310 0.8180804

head(louv_indig_big$modularity)  # [1] 0.6202836 0.7887157 0.8178573 0.8180804
tail(louv_indig_big$modularity)  # [1] 0.6202836 0.7887157 0.8178573 0.8180804

```


```{r 2d-big) examine modularity and degrees}

# this block is essentially copy-pasted from Liza's code

pref_seed <- seeds[which(mods == max(mods))]

set.seed(pref_seed)  # Liza comment:  # 4262 with min size of 80, 2925 with min size of 30

V(cit_graph_big_main)$louv_cluster <- louv_indig_big$membership

# modularity measure
    # B_ij = A_ij - γ k_i k_j / (2m)  == for UNDIRECTED graphs
    # B_ij = A_ij - γ k^out_i k^in_j / m == for DIRECTED graphs

modularity(louv_indig_big)
B = modularity_matrix(cit_graph_big_main)


# out-degree, in-degree, total-degree 

summary(degree(cit_graph_big_main,mode = 'out'))
  # Min.    1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 0.000    0.000   1.000   1.457   1.000  35.000 

summary(degree(cit_graph_big_main,mode = 'in'))
  #  Min.  1st Qu.   Median     Mean    3rd Qu.     Max. 
  #  0.000   0.000   1.000      1.457   1.000     76.000

summary(degree(cit_graph_big_main, mode = "all"))
  #  Min. 1st Qu.  Median    Mean   3rd Qu.    Max. 
  # 1.000   1.000   1.000   2.913   2.000     78.000 

# Add network summary
table(louv_indig_big$membership)
mean(table(louv_indig_big$membership))  # [1] 100.4333
median(table(louv_indig_big$membership))  # [1] 76
min(table(louv_indig_big$membership))  # [1] 12
max(table(louv_indig_big$membership))  # [1] 442

```


```{r e-big) PageRank measure of citation importance}

prank <- page_rank(cit_graph_big)
V(cit_graph_big)$prank <- unlist(as.vector(prank$vector))

betw <- betweenness(cit_graph_big)
V(cit_graph_big)$betweenness <- as.vector(betw)
betw_group <- data.table(node = V(cit_graph_big)$name,
                         between = betw,
                         group = V(cit_graph_big_main)$louv_cluster)
betw_group[,grank:=rank(-between,ties.method = 'average'),by=.(group)]

# when using "data.table" package, 
# don't have to re-store or re-assign variable
    # so the code above DOES create a column for "grank" even though it doesn't look like it
        # (which is why the next line is able to call it)

V(cit_graph_big)$high_between <- V(cit_graph_big)$name %in% betw_group[grank<=4,]$node

# Liza commented out the next line
# V(cit_graph)[betw == max(betw)]

prank_max <- data.table(node = V(cit_graph_big)$name,
                        group = unlist(V(cit_graph_big_main)$louv_cluster),
                        pagerank = unlist(V(cit_graph_big)$prank))

prank_max_by_group <- prank_max[,max(pagerank),by=.(group)]

max_by_group = dplyr::left_join(prank_max,prank_max_by_group)[pagerank==V1,]

V(cit_graph_big)$highest_in_group <- V(cit_graph_big)$name %in% max_by_group$node

saveRDS(cit_graph_big, "indig_cit_full.rds")
# indig_cit_full <- readRDS(file = "data/indig_cit_full.rds")

```

