---
title: "03a.1_networks_indig_open"
author: "Mackenzie Johnson"
date: "2026-10-02"
output: html_document
---

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
# knitr::opts_chunk$set(include = FALSE)

```

```{r libraries}

# install.packages("pbapply")
# install.packages("slam")
# install.packages("ggraph")

library(data.table)
library(ggplot2)
library(ggraph)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)
library(stringr)
library(tidyverse)

```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# edgelists

el_big <- readRDS(file = "data/edgelist_indig_open_20261001.rds")  # 4455 observations of 2 variables
# el_sm <- readRDS(file = "data/edgelist_indig_insular_20261001.rds")  # 724 observations of 2 variables

```

[UPDATE:  4/27/2026 and 5/12/2016

Started trying to make sense of the network again now that I have a new edgelist.  I brought the clustering algorithm out of the for-loop.  Made it through 2c.]


Step 1)  Create a citation graph following Wood et al. (2025) - code is available in Liza Wood's Github repository "tamar"
        a) start with a full graph
        b) simplify it by removing any multiple edges or loop edges - BUT!  WARNING!  Elise says "simplify" is notoriously wonky
        c) test if it's "connected"

        
Regarding "is_connected()"

From "networkx.org" documentation for doing network analysis in Python:
  * "A graph is connected if, for every pair of distinct nodes, there is a path between them. If there is a pair of nodes for which such path does not exist, the graph is not connected (also referred to as “disconnected”)"

From github:  https://olizardo.github.io/networks-textbook/lesson-graphs-connectivity.html
  * "A graph is CONNECTED if there exists a (finite) path between each pair of nodes.  A graph is DISCONNECTED if there does not exist a path connecting every pair of nodes"
  
From github:  https://yunranchen.github.io/intro-net-r/igraph.html
  * "Connected: if there is a path from any vertex to any other.  Unconnected: if not connected. An unconnected graph has multiple components." 
       
       
```{r 1-big) create initial citation graph}

# this block is essentially copy-pasted from Liza's code
  # only difference is I added some annotations

# this run is for articles with NAIS-themed titles or primary_topic
    # in the FULL list of works (after filtering/before limiting to the unique refs)
    # meaning: these works might reference works that do NOT have NAIS-themed titles or topics


cit_graph_indig_open <- graph_from_edgelist(as.matrix(el_big), directed = T)  # List of 3170

# str(cit_graph_big)


cit_graph_indig_open <- igraph::simplify(cit_graph_indig_open, remove.multiple = T, remove.loops = T)
  # remove.multiple	Logical, whether the multiple edges are to be removed
  # remove.loops	Logical, whether the loop edges are to be removed


# Liza didn't do this next step, but I want to check

is_connected(cit_graph_indig_open, mode = "weak")  # [1] FALSE

# FALSE means that it is disconnected (meaning, there are gaps between some of the nodes)

count_components(cit_graph_indig_full)  # [1] 25 --- matches component details in the next chunk

# to get a histogram for the maximal connected component sizes: 
  # component_distribution(cit_graph_indig_full)

```


Step 2) Network components ("maximal induced subgraph that is connected")
        a) Find connected components of the citation graph and visualize the network
        b) For those components from the main cluster (with 69,041 elements out of 69,144), create subgraph

[In the next workflow segment:]
Step 3) a) run clustering algorithm on the main cluster - start with Louvain
        b) examine modularity
        c) examine Page Rank estimates of citation importance
    

```{r 2a-big) find connected components of the larger graph}

# this will be easier if I rename the citation graph

cit_graph <- cit_graph_indig_open

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "comps"
  # added some annotations

# despite the "FALSE" result above for "is_connected" 
# ran the "components" function using both "weak" and "strong" modes
    # the "weak" run yielded the same thing as not specifying the mode, 
    # the "strong" run yielded 3250 clusters - seems like a Directed Acyclic Graph (?)
 

comps_indig_open <- components(cit_graph, mode = "weak")  # List of 3: $ membership, $ csize, $ no
str(comps_indig_open)
  # List of 3
  # $ membership: Named num [1:3170] 1 1 1 1 1 1 1 1 1 1 ...
    # ..- attr(*, "names")= chr [1:3170] "W1579625741" "W1995811201" "W2123089587" "W150613307"  ...
  # $ csize     : num [1:25] 2936 15 14 10 6 ..
  # $ no        : num 25

# --- grouping/summarizing doesn't work --- #

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(comps_indig_open$membership)

#   1     2    3    4    5    6    7    8    9   10   11   12   13   14   15   16   17   18   19   20   21   22 
# 2936   15   14   10    6    7   10    6    7    8   11    9   11   11   12    6    6   17   18   12    9    7 
 
#  23   24   25 
#   9    7    6 

# translation:  there are 25 clusters - 2936 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_indig_open <- names(comps_indig_open$membership)[comps_indig_open$membership == 1]

str(connected_indig_open)  #  [1:2936]  
    # --> which means 234 works were NOT INCLUDED in the main cluster (= 3170-2936)

# Got some visualization code from Sara!  

ggraph(cit_graph, layout = "kk") +
geom_edge_fan(
end_cap = circle(1, "mm"),
color = "#000000",
width = 0.3,
arrow = arrow(angle = 15, length = unit(0.07, "inches"), ends = "last", type = "closed"))+
geom_node_point(alpha = 0.2) +
labs(title = "NAIS-themed Citations (open network)") + theme_void()

```


```{r 2b-big) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "comps" but no other changes

# the following code is for extracting the main blob, 
# but I also want to see what happens when I run it on the full network
    # especially given that it seems like there are multiple clusters when I plotted the network 

if(any(comps_indig_open$membership>1)){
  cit_graph_main <- induced_subgraph(cit_graph,v = which(V(cit_graph)$name %in% connected_indig_open))
  cit_graph <- cit_graph_main
}

cit_graph_ud <- as.undirected(cit_graph, 'collapse')


# save the citation graph network
saveRDS(cit_graph, "data/indig_cit_open.rds")

# save the undirected citation graph network for use in clustering
saveRDS(cit_graph_ud, "data/indig_cit_open_ud.rds")

```

