---
title: "04_cluster_descrip_indig_full_20260512"
author: "Mackenzie Johnson"
date: "2026-05-12"
output: html_document
---

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
# knitr::opts_chunk$set(include = FALSE)

```

```{r libraries}

# install.packages("pbapply")
# install.packages("slam")
# install.packages("ggraph")
# install.packages("udpipe")  
      # Tokenising, Lemmatising, Tagging and Dependency Parsing of raw text in TIF format

library(data.table)
library(dplyr)
library(ggplot2)
library(ggraph)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)  # Sparse Lightweight Arrays and Matrices
library(stringr)
library(tidyverse)
library(udpipe)

```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# nodelist (aka "metadata")

M_dt <- readRDS(file = "nodelist_Indig_full.rds")

# citation graph - with data from cluster algorithm

cit_graph <- readRDS(file = "data/indig_clusters_full.rds")

dim(M_dt)  # [1] 3250   21
summary(cit_graph)  # List of 3013

# difference is the 237 "unconnected" components

```

To Recap:

I filtered the "works" to JUST those with NAIS-themed keywords in their title or primary_topic variables.  I found the "refs" that matched those work_ids - the resultant edgelist includes NAIS-themed works that appear in EITHER the "citing" ("from") or "cited" ("to") columns of the "refs" object.  That means they could be citing or cited by works that do NOT have NAIS-themed titles or primary-topics.  Using the edgelist, created a citation network and ran the Louvain clustering algorithm to identify clusters.  

The citation graph I'm starting with here "indig_clus_full" includes node modularity and betweenness centrality values, cluster membership, and 2 ranked measures of citation importance (page rank and grank).  

Here I want to link the citation graph to the nodelist and then explore the combined dataset: descriptive statistics and network visualizations that display some of the node attributes.


Step 1)  Add an attribute in the node-list for the cluster membership
       
       
```{r 1) add cluster membership to node-list}

# this block is essentially copy-pasted from Liza's code

M_dt$cluster <- V(cit_graph)$louv_cluster[match(M_dt$id,V(cit_graph)$name)]

table(M_dt$cluster)  # < table of extent 0 >

```


Step 2) Identify clusters
        a) Find connected components of the citation graph and visualize the network
        b) For those components from the main cluster (with 69,041 elements out of 69,144), create subgraph
        c) run clustering algorithm on the main cluster - start with Louvain
        d) examine modularity
        e) examine Page Rank estimates of citation importance
    

```{r 2a-big) find connected components of the larger graph}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus"
  # added some annotations

# despite the "FALSE" result above for "is_connected" 
# ran the "components" function using both "weak" and "strong" modes
# the "weak" run yielded the same thing as not specifiying the mode, 
# the "strong" run yielded 3250 clusters LOL!!!
 

clus_big <- components(cit_graph_big)  # List of 3: $ membership, $ csize, $ no
str(clus_big)
  # List of 3
  # $ membership: Named num [1:3250] 1 1 1 1 1 1 1 1 1 1 ...
    # ..- attr(*, "names")= chr [1:3250] "W1579625741" "W1995811201" ...
  # $ csize     : num [1:25] 3013 15 14 10 6 ... ...
  # $ no        : num 25

# --- grouping/summarizing doesn't work --- "

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(clus_big$membership)

# 1       2    3    4    5    6    7    8    9   10   11   12   13   14   15   16   17   18   19 
# 3013   15   14   10    6    7   10    6    7   10   12    9   11   11   12    6    6   17   18 

#   20   21   22   23   24   25 
#   12    9    7    9    7    6 

# translation:  there are 25 clusters - 3013 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_component_big <- names(clus_big$membership)[clus_big$membership == 1]

str(connected_component_big)  #  [1:3013]  
    # --> which means 237 works were NOT INCLUDED in the main cluster (3250-3013)

# Got some visualization code from Sara!  

ggraph(cit_graph_big, layout = "fr") +
geom_edge_fan(
end_cap = circle(1, "mm"),
color = "#000000",
width = 0.3,
arrow = arrow(angle = 15, length = unit(0.07, "inches"), ends = "last", type = "closed"))+
geom_node_point(alpha = 0.2) +
labs(title = "NAIS-themed Citations (full works)") + theme_void()

```


Before moving on, I tried figuring out a for-loop to get the cluster membership for each work_id.

It didn't go well...need to come back to this.

unconnecteds_names <- names(clus$membership)[clus$membership != 1]  #  chr [1:1729]
    # can eventually use this list to find the "outliers"
    # as Liza did

clus_names <- names(clus$membership)[clus$membership != 1]
clus_num <- clus$no

cluster_list <- clus$membership



cluster_assign <- for(i in cluster_list) {
  which(clus$membership == i)
}

length(clus$membership)


```{r 2b-big) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus" but no other changes

if(any(clus_big$membership>1)){
  cit_graph_big_main <- induced_subgraph(cit_graph_big,v = which(V(cit_graph_big)$name %in% connected_component_big))
  cit_graph_big <- cit_graph_big_main
}

cit_graph_big_ud <- as.undirected(cit_graph_big, 'collapse')

```


```{r 2c-big) run clustering algorithm}

# this block is essentially copy-pasted from Liza's code

# recall that she had a much higher seed value under certain call conditions
      # if(LOCATION == 'data/query_EG/' & ADD_SUPPL_WOS_REFS == F){
        #  pref_seed <- 4812
        #  } else{

# will try running it with the set.seed set to 12 
    # idea is to seed a pseudo random number generator using a reproducible seed 
    # select 1000 numbers with equal weighting from the set 0:9999
    # why do we need this?

set.seed(12)  # Liza comment: # and 16 got to 8, dropped PES and joined it with PE
seeds <- sample.int(9999, 1000)

mod_big <- list()
for(i in 1:length(seeds)){
  set.seed(seeds[i])
      # why is the clustering algo in the for-loop? 
      # Not affected by iteration, move to outside of loop for efficiency
  louv_indig_big <- cluster_louvain(cit_graph_big_ud)
  mod_big$mod[i] <- modularity(louv_indig_big)
  mod_big$min_n[i] <- min(table(louv_indig_big$membership))
}
mods <- unlist(mod_big$mod)
min_ns <- unlist(mod_big$min_n)

# saveRDS(louv_indig, "Louvain_indig")
# louv <- readRDS(file = "data/Louvain_full")
    
head(mods)  # [1] 0.8170753 0.8162952 0.8183118 0.8163955 0.8195041 0.8157186
tail(mods)  # [1] 0.8154065 0.8164846 0.8171450 0.8147561 0.8165310 0.8180804

head(louv_indig_big$modularity)  # [1] 0.6202836 0.7887157 0.8178573 0.8180804
tail(louv_indig_big$modularity)  # [1] 0.6202836 0.7887157 0.8178573 0.8180804

```


```{r 2d-big) examine modularity and degrees}

# this block is essentially copy-pasted from Liza's code

pref_seed <- seeds[which(mods == max(mods))]

set.seed(pref_seed)  # Liza comment:  # 4262 with min size of 80, 2925 with min size of 30

V(cit_graph_big_main)$louv_cluster <- louv_indig_big$membership

# modularity measure
    # B_ij = A_ij - γ k_i k_j / (2m)  == for UNDIRECTED graphs
    # B_ij = A_ij - γ k^out_i k^in_j / m == for DIRECTED graphs

modularity(louv_indig_big)
B = modularity_matrix(cit_graph_big_main)


# out-degree, in-degree, total-degree 

summary(degree(cit_graph_big_main,mode = 'out'))
  # Min.    1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 0.000    0.000   1.000   1.457   1.000  35.000 

summary(degree(cit_graph_big_main,mode = 'in'))
  #  Min.  1st Qu.   Median     Mean    3rd Qu.     Max. 
  #  0.000   0.000   1.000      1.457   1.000     76.000

summary(degree(cit_graph_big_main, mode = "all"))
  #  Min. 1st Qu.  Median    Mean   3rd Qu.    Max. 
  # 1.000   1.000   1.000   2.913   2.000     78.000 

# Add network summary
table(louv_indig_big$membership)
mean(table(louv_indig_big$membership))  # [1] 100.4333
median(table(louv_indig_big$membership))  # [1] 76
min(table(louv_indig_big$membership))  # [1] 12
max(table(louv_indig_big$membership))  # [1] 442

```


```{r e-big) PageRank measure of citation importance}

prank <- page_rank(cit_graph_big)
V(cit_graph_big)$prank <- unlist(as.vector(prank$vector))

betw <- betweenness(cit_graph_big)
V(cit_graph_big)$betweenness <- as.vector(betw)
betw_group <- data.table(node = V(cit_graph_big)$name,
                         between = betw,
                         group = V(cit_graph_big_main)$louv_cluster)
betw_group[,grank:=rank(-between,ties.method = 'average'),by=.(group)]


V(cit_graph_big)$high_between <- V(cit_graph_big)$name %in% betw_group[grank<=4,]$node

# Liza commented out the next line
# V(cit_graph)[betw == max(betw)]

prank_max <- data.table(node = V(cit_graph_big)$name,group = unlist(V(cit_graph_big_main)$louv_cluster),pagerank = unlist(V(cit_graph_big)$prank))
prank_max_by_group <- prank_max[,max(pagerank),by=.(group)]

max_by_group = dplyr::left_join(prank_max,prank_max_by_group)[pagerank==V1,]

V(cit_graph_big)$highest_in_group <- V(cit_graph_big)$name %in% max_by_group$node

saveRDS(cit_graph_big, "indig_cit_full.rds")
# indig_cit_full <- readRDS(file = "data/indig_cit_full.rds")

```

