---
title: "04_cluster_descrip_indig_full_20260512"
author: "Mackenzie Johnson"
date: "2026-05-12"
output: html_document
---

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
# knitr::opts_chunk$set(include = FALSE)

```

```{r libraries}

install.packages(c("visNetwork", 
                   "sfnetworks", 
                   "tidygraph"))

# install.packages("ggraph")
# install.packages("udpipe")  
      # Tokenising, Lemmatising, Tagging and Dependency Parsing of raw text in TIF format

library(data.table)
library(dplyr)
library(ggplot2)
library(ggraph)
library(igraph)
library(knitr)
# library(Matrix)
# library(pbapply)
library(sfnetworks)
# library(slam)  # Sparse Lightweight Arrays and Matrices
library(stringr)
library(tidygraph)
library(tidyverse)
library(udpipe)
library(visNetwork)

```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# nodelist (aka "metadata")

M_dt <- readRDS(file = "nodelist_Indig_full.rds")

# citation graph - with data from cluster algorithm

cit_graph <- readRDS(file = "data/indig_clusters_full.rds")

dim(M_dt)  # [1] 3250   21
summary(cit_graph)  # List of 3013

# difference is the 237 "unconnected" components

```

To Recap:

I filtered the "works" to JUST those with NAIS-themed keywords in their title or primary_topic variables.  I found the "refs" that matched those work_ids - the resultant edgelist includes NAIS-themed works that appear in EITHER the "citing" ("from") or "cited" ("to") columns of the "refs" object.  That means they could be citing or cited by works that do NOT have NAIS-themed titles or primary-topics.  Using the edgelist, created a citation network and ran the Louvain clustering algorithm to identify clusters.  

The citation graph I'm starting with here "indig_clus_full" includes node modularity and betweenness centrality values, cluster membership, and 2 ranked measures of citation importance (page rank and grank).  

Here I want to link the citation graph to the nodelist and then explore the combined dataset: descriptive statistics and network visualizations that display some of the node attributes.


Step 1)  Add attributes in the node-list for the network characteristics:  cluster membership, page rank, and betweenness metrics.
       
       
```{r 1) add network attributes to node-list}

# this line is essentially copy-pasted from Liza's code
# adds cluster number to each work

M_dt$cluster <- V(cit_graph)$louv_cluster[match(M_dt$work_id,V(cit_graph)$name)]

table(M_dt$cluster)  

# also want to find the works with the highest page rank and betweenness metrics

M_dt$prank <- V(cit_graph)$prank[match(M_dt$work_id,V(cit_graph)$name)]
M_dt$btw <- V(cit_graph)$betweenness[match(M_dt$work_id,V(cit_graph)$name)]
M_dt$hi_btw <- V(cit_graph)$high_between[match(M_dt$work_id,V(cit_graph)$name)]
M_dt$highest <- V(cit_graph)$highest_in_group[match(M_dt$work_id,V(cit_graph)$name)]

# really don't need some of these original columns

M_dt_new <- M_dt %>% 
  select(-has_fulltext, -oa_is_oa, -oa_status)

# class(M_dt_new)  # [1] "data.table" "data.frame"
# typeof(M_dt_new)  # [1] "list"

saveRDS(M_dt_new, "indig_full_nodelist.rds")
# readRDS(file = "data/indig_full_nodelist.rds")

```


Step 2) Explore the new node-list.
  
  * overall distribution of articles between the 3 sources
  * group by cluster and summarize to compare average page rank and betweenness
  * for each cluster, look at the primary_topic variables ==> FOR LOOP!!


```{r 2) exploring the new node list}




```


 Step 3) extracting information from the new node-list

  * find the papers with the highest page-rank and betweenness metrics
  * use "highest" variable to find the works in each group with the highest page-rank values
  

```{r 3) extracting specific works based on network}


```


Step 4) visualize the network and use different node attributes

```{r 4) network visualization}

# Got some visualization code from Sara!  

ggraph(cit_graph_big, layout = "fr") +
geom_edge_fan(
end_cap = circle(1, "mm"),
color = "#000000",
width = 0.3,
arrow = arrow(angle = 15, length = unit(0.07, "inches"), ends = "last", type = "closed"))+
geom_node_point(alpha = 0.2) +
labs(title = "NAIS-themed Citations (full works)") + theme_void()

```

