---
title: "03_bibliometrics_networks"
author: "Mackenzie Johnson"
date: "2026-04-09"
output: html_document
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r libraries}
# install.packages("pbapply")
# install.packages("slam")

library(data.table)
library(ggplot2)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)
library(stringr)
library(tidyverse)

```


```{r code for writing csvs}

output_directory <- "C:/Users/Owner/Documents/Research/R_Projects/nais_ps/data_output/"
file_name_fields <- "Subfields_Topics.csv"
full_output_path_fields <- file.path(output_directory, file_name_fields)

write_excel_csv(df, file = full_output_path_fields)

```



```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# raw data

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

# edgelist

el_indig <- readRDS(file = "data/network_edgelist_indig")

```

Step 1)  Create a citation graph following Wood et al. (2025) - code is available in Liza Wood's Github repository "tamar"
        a) start with a full graph
        b) simplify it by removing any multiple edges or loop edges - BUT!  WARNING!  Elise says "simplify" is notoriously wonky

```{r 1) create citation graph}

# this block is essentially copy-pasted from Liza's code
  # only difference is I added some annotations

cit_graph <- graph_from_edgelist(as.matrix(el_indig), directed = T)

# str(cit_graph)

# Class 'igraph'  hidden list of 10
#  $ : num 69144

# result is a "Large igraph (69144 elements, 12.4 MB)

cit_graph <- igraph::simplify(cit_graph, remove.multiple = T, remove.loops = T)
  # remove.multiple	Logical, whether the multiple edges are to be removed
  # remove.loops	Logical, whether the loop edges are to be removed

# result is a "Large igraph (69144 elements, 5 MB)

```


Step 2) Identify clusters
        a) Find connected components of the citation graph
        b) For those components from the main cluster (with 69,041 elements out of 69,144), create subgraph
        c) run clustering algorithm on the main cluster - start with Louvain
        d) examine modularity
        e) delve into the clusters
    

```{r 2a) find connected components}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus"
  # added some annotations

clus <- components(cit_graph)  # List of 3: $ membership, $ csize, $ no

str(clus)
  # List of 3
  # $ membership: Named num [1:990] 1 1 1 1 1 1 1 1 1 1
    # ..- attr(*, "names")= chr [1:990] "W2599301038" ...
  # $ csize     : num [1:5] 913 9 37 19 12
  # $ no        : num 5

# --- grouping/summarizing doesn't work --- "

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(clus$membership)

#   1   2   3   4   5       
# 913   9  37  19  12 

# translation:  there are 5 clusters - 913 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_component <- names(clus$membership)[clus$membership == 1]

str(connected_component)  #  chr [1:913]  
    # --> which means 77 works were NOT INCLUDED in the main cluster (= 990-913)

unconnecteds <- names(clus$membership)[clus$membership != 1]  #  chr [1:103]
    # can eventually use this list to find the outliers
    # as Liza did

```

```{r 2b) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus" but no other changes

if(any(clus$membership>1)){
  cit_graph_main <- induced_subgraph(cit_graph,v = which(V(cit_graph)$name %in% connected_component))
  # cit_graph <- cit_graph_main  # Liza recycled the name "cit_graph" but I opted to keep it as "..._main"
}

cit_graph_ud <- as.undirected(cit_graph_main, 'collapse')

```


```{r 2c) run clustering algorithm}

# recall that she had a much higher seed value under certain call conditions
      # if(LOCATION == 'data/query_EG/' & ADD_SUPPL_WOS_REFS == F){
        #  pref_seed <- 4812
        #  } else{

# will try running it with the set.seed set to 12 

# trying to run the clustering algorithm and 
# subsequent modularity metrics with the clustering outside the for-loop

louv_indig <- cluster_louvain(cit_graph_ud)
 set.seed(12)  # Liza comment: # and 16 got to 8, dropped PES and joined it wit PE
    seeds <- sample.int(9999, 1000)
    # louv <- cluster_louvain(cit_graph_ud)
    mod <- list()
    for(i in 1:length(seeds)){
      set.seed(seeds[i])
      
      mod$mod[i] <- modularity(louv_indig)
      mod$min_n[i] <- min(table(louv_indig$membership))
    }
    mods <- unlist(mod$mod)
    min_ns <- unlist(mod$min_n)
    
# time to run:  ~52 minutes
# per Elise's guidance, should probably save the rds!
    
saveRDS(louv, "Louvain_indig")
louv <- readRDS(file = "data/Louvain")
    
head(mods)  # [1] 0.6961121 0.6974466 0.6971936 0.6956488 0.6984532 0.6987537
tail(mods)  # [1] 0.6989354 0.6966139 0.6955635 0.6962756 0.6958860 0.6961305


```

```{r 2d) examine modularity and degrees}

pref_seed <- seeds[which(mods == max(mods))]

set.seed(pref_seed)  # Liza comment:  # 4262 with min size of 80, 2925 with min size of 30
louv <- readRDS(file = "data/Louvain")
V(cit_graph_main)$louv_cluster <- louv$membership

# modularity measure

modularity(louv)
B = modularity_matrix(cit_graph_main)

    # Error in modularity_matrix(cit_graph_main) :
        # At vendor/cigraph/src/core/vector.pmt:475 : Cannot reserve space for vector. Out of memory

# out-degree, in-degree, total-degree 

summary(degree(cit_graph_main,mode = 'out'))
  # Min.    1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 0.000   1.000   4.000   6.673  10.000   190.000

summary(degree(cit_graph_main,mode = 'in'))
  #  Min.   1st Qu.  Median    Mean 3rd Qu.    Max. 
  #  0.000   0.000   1.000   6.673   7.000   791.000

summary(degree(cit_graph_main, mode = "all"))
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 1.00    2.00    6.00   13.35   18.00    796.00

# Add network summary
table(louv_indig$membership)
mean(table(louv$membership))  # [1] 3138.227
median(table(louv$membership))  # [1] 1482.5
min(table(louv$membership))  # [1] 10
max(table(louv$membership))  # [1] 14097

```



```{r e) cluster descriptions}

# not sure "page_rank" is the best approach

prank <- page_rank(cit_graph)
V(cit_graph)$prank <- unlist(as.vector(prank$vector))

betw <- betweenness(cit_graph)
V(cit_graph)$betweenness <- as.vector(betw)
betw_group <- data.table(node = V(cit_graph)$name,between = betw,group = V(cit_graph)$louv_cluster)
betw_group[,grank:=rank(-between,ties.method = 'average'),by=.(group)]


V(cit_graph)$high_between <- V(cit_graph)$name %in% betw_group[grank<=4,]$node

#V(cit_graph)[betw == max(betw)]
prank_max <- data.table(node = V(cit_graph)$name,group = unlist(V(cit_graph)$louv_cluster),pagerank = unlist(V(cit_graph)$prank))
prank_max_by_group <- prank_max[,max(pagerank),by=.(group)]

max_by_group = dplyr::left_join(prank_max,prank_max_by_group)[pagerank==V1,]

V(cit_graph)$highest_in_group <- V(cit_graph)$name %in% max_by_group$node

saveRDS(cit_graph, paste0(path_location,'cit_graph.rds'))

```



--- Old notes --- 
(from when the louvain clustering algorithm did NOT work correctly, 
and when I totally misunderstood what the "clus" table was telling me...)

Want to go back at this point and see if setting up the connnected components like Liza did makes a difference.  Her approach was to construct the "connected_components" with values equal to 1,then do the same "if > 1" to make the subgraph.  

Given that she linked her "connected_components" with the metadata to get the titles of those with membership == 1, she might have set up her "connected components" in this way because she had a lot fewer components with only 1 edge.  However, in the NAIS-PS data the majority have a membership of 1:

    #     1     2     3     4     5     6     7     8     9 
    # 69041     9    11    19    15     9    15    12    13 
    

After re-doing steps 2a-b using her approach (see below), the objects produced have the same dimensions and same head() and tail() values as those objects I produced using my approach.  

And the "louv" communities object seems broken - might have been because I had made it using the outlying clusters instead of the main cluster.  

  - when I try to view "louv" - get an error message:
      
        # Error in names[[i]] : subscript out of bounds
        
   - Global environment:  "louv" is a List of 8
   - can't view it
   - CAN see six variables when I click on the drop-down (but shouldn't it be 8?)
   - louv[[1]] - gives me 9 work_id values ... louv[[2]] give me 11 values ...
   - in other words, the number of work_id values provided when calling each column matches the number in the membership summary table (see above) 


