---
title: "03_bibliometrics_networks"
author: "Mackenzie Johnson"
date: "2026-04-09"
output: html_document
---

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
# knitr::opts_chunk$set(include = FALSE)

```

```{r libraries}

# install.packages("pbapply")
# install.packages("slam")
# install.packages("ggraph")

library(data.table)
library(ggplot2)
library(ggraph)
library(igraph)
library(knitr)
library(Matrix)
library(pbapply)
library(slam)
library(stringr)
library(tidyverse)

```


```{r SetWD}

getwd()  # [1] "C:/Users/Owner/Documents/Research/R_Projects/nais_ps"
# setwd() 
```


```{r GetData}

# raw data

works <- readRDS(file = "data/works.rds")
refs <- readRDS(file = "data/refs.rds")

# edgelist

el_big <- readRDS(file = "data/edgelist_indig_or")  # 4603 observations of 2 variables
el_sm <- readRDS(file = "data/edgelist_indig_and")  # 727 observations of 2 variables

```

[UPDATE:  4/27/2026

Started trying to make sense of the network again now that I have a new edgelist.  I brought the clustering algorithm out of the for-loop.  Made it through 2c.]


Step 1)  Create a citation graph following Wood et al. (2025) - code is available in Liza Wood's Github repository "tamar"
        a) start with a full graph
        b) simplify it by removing any multiple edges or loop edges - BUT!  WARNING!  Elise says "simplify" is notoriously wonky

```{r 1) create citation graphs}

# this block is essentially copy-pasted from Liza's code
  # only difference is I added some annotations

cit_graph_big <- graph_from_edgelist(as.matrix(el_big), directed = T)  # List of 3250

# str(cit_graph_big)


cit_graph_big <- igraph::simplify(cit_graph_big, remove.multiple = T, remove.loops = T)
  # remove.multiple	Logical, whether the multiple edges are to be removed
  # remove.loops	Logical, whether the loop edges are to be removed

# Liza didn't do this next step, but I want to check

is_connected(cit_graph_big, mode = "weak")  # [1] FALSE

# OK...what about the smaller network 
    # articles with NAIS-themed titles or primary_topic 
    # AND appear in the "unique" list of refs

cit_graph_sm <- graph_from_edgelist(as.matrix(el_sm), directed = T)   # List of 390
cit_graph_sm <- igraph::simplify(cit_graph_sm, remove.multiple = T, remove.loops = T)
is_connected(cit_graph_big, mode = "weak")  # [1] FALSE



```


Step 2) Identify clusters
        a) Find connected components of the citation graph
        b) For those components from the main cluster (with 69,041 elements out of 69,144), create subgraph
        c) run clustering algorithm on the main cluster - start with Louvain
        d) examine modularity
        e) examine Page Rank estimates of citation importance
    

```{r 2a-big) find connected components of the larger graph}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus"
  # added some annotations

# despite the "FALSE" result above for "is_connected" 
# ran the "components" function using both "weak" and "strong" modes
# the "weak" run yielded the same thing as not specifiying the mode, 
# the "strong" run yielded 3250 clusters LOL!!!
 

clus_big <- components(cit_graph_big)  # List of 3: $ membership, $ csize, $ no
str(clus_big)
  # List of 3
  # $ membership: Named num [1:3250] 1 1 1 1 1 1 1 1 1 1 ...
    # ..- attr(*, "names")= chr [1:3250] "W1579625741" "W1995811201" ...
  # $ csize     : num [1:25] 3013 15 14 10 6 ... ...
  # $ no        : num 25

# --- grouping/summarizing doesn't work --- "

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(clus_big$membership)

# 1       2    3    4    5    6    7    8    9   10   11   12   13   14   15   16   17   18   19 
# 3013   15   14   10    6    7   10    6    7   10   12    9   11   11   12    6    6   17   18 

#   20   21   22   23   24   25 
#   12    9    7    9    7    6 

# translation:  there are 25 clusters - 3013 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_component_big <- names(clus_big$membership)[clus_big$membership == 1]

str(connected_component_big)  #  [1:3013]  
    # --> which means 237 works were NOT INCLUDED in the main cluster (3250-3013)

```

```{r 2a-sm) find connected components of the smaller (unique refs) graph}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus"
  # added some annotations

# despite the "FALSE" result above for "is_connected" 
# ran the "components" function using both "weak" and "strong" modes
# the "weak" run yielded the same thing as not specifiying the mode, 
# the "strong" run yielded 3250 clusters LOL!!!
 

clus_sm <- components(cit_graph_sm)  # List of 3: $ membership, $ csize, $ no
str(clus_sm)
  # List of 3
  # $ membership: Named num [1:3390] 1 1 1 1 1 1 1 1 1 1 ...
    # ..- attr(*, "names")= chr [1:3250] "W1579625741" "W1995811201" ...
  # $ csize     : num [1:12] 358 2 3 2 2 2 4 2 4 7 ... ...
  # $ no        : num 12

# --- grouping/summarizing doesn't work --- "

    # Error in UseMethod("group_by") : 
        # no applicable method for 'group_by' applied to an object of class "list"


table(clus_sm$membership)

#   1   2   3   4   5   6   7   8   9  10  11  12 
# 358   2   3   2   2   2   4   2   4   7   2   2 

# translation:  there are 12 clusters - 358 elements are all in 1 cluster
# meaning, need to compute the communities JUST using the elements in the main cluster

connected_component_sm <- names(clus_sm$membership)[clus_sm$membership == 1]

str(connected_component_sm)  #  [1:3013]  
    # --> which means 32 works were NOT INCLUDED in the main cluster (390-358)

```


Before moving on, I tried figuring out a for-loop to get the cluster membership for each work_id.

It didn't go well...need to come back to this.

unconnecteds_names <- names(clus$membership)[clus$membership != 1]  #  chr [1:1729]
    # can eventually use this list to find the "outliers"
    # as Liza did

clus_names <- names(clus$membership)[clus$membership != 1]
clus_num <- clus$no

cluster_list <- clus$membership



cluster_assign <- for(i in cluster_list) {
  which(clus$membership == i)
}

length(clus$membership)


```{r 2b-big) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus" but no other changes

if(any(clus_big$membership>1)){
  cit_graph_big_main <- induced_subgraph(cit_graph_big,v = which(V(cit_graph_big)$name %in% connected_component_big))
  cit_graph_big <- cit_graph_big_main
}

cit_graph_big_ud <- as.undirected(cit_graph_big, 'collapse')

```


```{r 2b-sm) create subgraph and undirected version}

# this block is essentially copy-pasted from Liza's code
  # changed "clu" to "clus" but no other changes

if(any(clus_sm$membership>1)){
  cit_graph_sm_main <- induced_subgraph(cit_graph_sm,v = which(V(cit_graph_sm)$name %in% connected_component_sm))
  cit_graph_sm <- cit_graph_sm_main
}

cit_graph_sm_ud <- as.undirected(cit_graph_sm, 'collapse')

```


```{r 2c-big) run clustering algorithm}

# this block is essentially copy-pasted from Liza's code

# recall that she had a much higher seed value under certain call conditions
      # if(LOCATION == 'data/query_EG/' & ADD_SUPPL_WOS_REFS == F){
        #  pref_seed <- 4812
        #  } else{

# will try running it with the set.seed set to 12 
    # idea is to seed a pseudo random number generator using a reproducible seed 
    # select 1000 numbers with equal weighting from the set 0:9999
    # why do we need this?

set.seed(12)  # Liza comment: # and 16 got to 8, dropped PES and joined it with PE
seeds <- sample.int(9999, 1000)

mod_big <- list()
for(i in 1:length(seeds)){
  set.seed(seeds[i])
      # why is the clustering algo in the for-loop? 
      # Not affected by iteration, move to outside of loop for efficiency
  louv_indig_big <- cluster_louvain(cit_graph_big_ud)
  mod_big$mod[i] <- modularity(louv_indig_big)
  mod_big$min_n[i] <- min(table(louv_indig_big$membership))
}
mods <- unlist(mod_big$mod)
min_ns <- unlist(mod_big$min_n)

# saveRDS(louv_indig, "Louvain_indig")
# louv <- readRDS(file = "data/Louvain_full")
    
head(mods)  # [1] 0.8170753 0.8162952 0.8183118 0.8163955 0.8195041 0.8157186
tail(mods)  # [1] 0.8154065 0.8164846 0.8171450 0.8147561 0.8165310 0.8180804

head(louv_indig_big$modularity)  # [1] 0.6202836 0.7887157 0.8178573 0.8180804
tail(louv_indig_big$modularity)  # [1] 0.6202836 0.7887157 0.8178573 0.8180804

```


```{r 2c-sm) run clustering algorithm}

# this block is essentially copy-pasted from Liza's code

# recall that she had a much higher seed value under certain call conditions
      # if(LOCATION == 'data/query_EG/' & ADD_SUPPL_WOS_REFS == F){
        #  pref_seed <- 4812
        #  } else{

# will try running it with the set.seed set to 12 
    # idea is to seed a pseudo random number generator using a reproducible seed 
    # select 1000 numbers with equal weighting from the set 0:9999
    # why do we need this?

set.seed(12)  # Liza comment: # and 16 got to 8, dropped PES and joined it with PE
seeds <- sample.int(9999, 1000)

mod_sm <- list()
for(i in 1:length(seeds)){
  set.seed(seeds[i])
      # why is the clustering algo in the for-loop? 
      # Not affected by iteration, move to outside of loop for efficiency
  louv_indig_sm <- cluster_louvain(cit_graph_sm_ud)
  mod_sm$mod[i] <- modularity(louv_indig_sm)
  mod_sm$min_n[i] <- min(table(louv_indig_sm$membership))
}
mods_sm <- unlist(mod_sm$mod)
min_ns_sm <- unlist(mod_sm$min_n)

# saveRDS(louv_indig, "Louvain_indig")
# louv <- readRDS(file = "data/Louvain_full")
    
head(mods_sm)  # [1] 0.7305574 0.7280482 0.7294993 0.7300628 0.7310972 0.7317859
tail(mods_sm)  # [1] 0.7247867 0.7298154 0.7307565 0.7296923 0.7297662 0.7339800

head(louv_indig_sm$modularity)  # [[1] 0.5736273 0.7130278 0.7337727 0.7339800
tail(louv_indig_sm$modularity)  # [1] 0.5736273 0.7130278 0.7337727 0.7339800

```


```{r 2d) examine modularity and degrees}

# this block is essentially copy-pasted from Liza's code

pref_seed <- seeds[which(mods == max(mods))]

set.seed(pref_seed)  # Liza comment:  # 4262 with min size of 80, 2925 with min size of 30
# louv <- readRDS(file = "data/Louvain_soc_env")
V(cit_graph_big_main)$louv_cluster <- louv_indig_big$membership

# modularity measure
    # B_ij = A_ij - γ k_i k_j / (2m)  == for UNDIRECTED graphs
    # B_ij = A_ij - γ k^out_i k^in_j / m == for DIRECTED graphs

modularity(louv_indig_big)
B = modularity_matrix(cit_graph_big_main)


# out-degree, in-degree, total-degree 

summary(degree(cit_graph_big_main,mode = 'out'))
  # Min.    1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 0.000    0.000   1.000   1.457   1.000  35.000 

summary(degree(cit_graph_big_main,mode = 'in'))
  #  Min.  1st Qu.   Median     Mean    3rd Qu.     Max. 
  #  0.000   0.000   1.000      1.457   1.000     76.000

summary(degree(cit_graph_big_main, mode = "all"))
  #  Min. 1st Qu.  Median    Mean   3rd Qu.    Max. 
  # 1.000   1.000   1.000   2.913   2.000     78.000 

# Add network summary
table(louv_indig_big$membership)
mean(table(louv_indig_big$membership))  # [1] 100.4333
median(table(louv_indig_big$membership))  # [1] 76
min(table(louv_indig_big$membership))  # [1] 12
max(table(louv_indig_big$membership))  # [1] 442

```



```{r e) PageRank measure of citation importance}

prank <- page_rank(cit_graph_big)
V(cit_graph_big)$prank <- unlist(as.vector(prank$vector))

betw <- betweenness(cit_graph_big)
V(cit_graph_big)$betweenness <- as.vector(betw)
betw_group <- data.table(node = V(cit_graph_big)$name,
                         between = betw,
                         group = V(cit_graph_big_main)$louv_cluster)
betw_group[,grank:=rank(-between,ties.method = 'average'),by=.(group)]


V(cit_graph_big)$high_between <- V(cit_graph_big)$name %in% betw_group[grank<=4,]$node

# Liza commented out the next line
# V(cit_graph)[betw == max(betw)]

prank_max <- data.table(node = V(cit_graph_big)$name,group = unlist(V(cit_graph_big_main)$louv_cluster),pagerank = unlist(V(cit_graph_big)$prank))
prank_max_by_group <- prank_max[,max(pagerank),by=.(group)]

max_by_group = dplyr::left_join(prank_max,prank_max_by_group)[pagerank==V1,]

V(cit_graph_big)$highest_in_group <- V(cit_graph_big)$name %in% max_by_group$node

saveRDS(cit_graph_big, "cit_graph_big.rds")
indig_cit_full <- readRDS(file = "data/cit_graph_big.rds")

```

